---
title: "SomaScan analysis of PD dementia"
author: "Raquel Real"
date: "27/06/2026"
---

# Set working directory
```{r}
#setwd("/path/to/working/directory")
setwd("/Users/raquelreal/Documents/Manuscripts/SomaScan/Test_code/")
```

# Load libraries
```{r}
library(SomaDataIO)
library(tidyverse)
library(tidyr)
library(dplyr)
library(data.table)
library(ggplot2)
library(ggrepel)
library(survival)
library(Biobase)
library(limma)
library(sva)
library(metafor)
library(broom)
library(CMAverse)
library(UpSetR)
library(clusterProfiler)
library(org.Hs.eg.db)
library(ReactomePA)
library(stringr)
library(STRINGdb)
library(GOSemSim)
```

# Load data
```{r}
#Discovery cohort
# load the SomaScan data 
soma_data_discovery <- read_adat("data/soma_data.adat", debug = FALSE, verbose = getOption("verbose"))

# load the sample clinical data
clinical_data_discovery <- fread("data/clinical_data.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)


#Replication cohort
# load the SomaScan data 
soma_data_replication <- read_adat("data/soma_data.adat", debug = FALSE, verbose = getOption("verbose"))

# load the sample clinical data
clinical_data_replication <- fread("data/clinical_data.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
```

# Clean the SomaScan data
```{r}
#Discovery cohort
# remove the rows (samples)that have failed the quality control
soma_data_discovery_PASS <- soma_data_discovery %>% filter(RowCheck == "PASS")

# remove the non-human proteins
AptamerNonHuman <- attributes(soma_data_discovery_PASS)$Col.Meta %>% filter(Organism != "Human" | Type != "Protein") %>% dplyr::select(SeqId)
AptamerNonHuman <- as.data.frame(AptamerNonHuman)
AptamerNonHuman$SeqId = paste0('seq.', AptamerNonHuman$SeqId)
AptamerNonHuman$SeqId <- sub('-', '.', AptamerNonHuman$SeqId, fixed = TRUE)

soma_data_discovery_PASS = soma_data_discovery_PASS %>% dplyr::select(!AptamerNonHuman[,1])


#Replication cohort
# remove the rows (samples)that have failed the quality control
soma_data_replication_PASS <- soma_data_replication %>% filter(RowCheck == "PASS")          

# remove the non-human proteins     
AptamerNonHuman <- attributes(soma_data_replication_PASS)$Col.Meta %>% filter(Organism != "Human" | Type != "Protein") %>% dplyr::select(SeqId)
AptamerNonHuman <- as.data.frame(AptamerNonHuman)
AptamerNonHuman$SeqId = paste0('seq.', AptamerNonHuman$SeqId)
AptamerNonHuman$SeqId <- sub('-', '.', AptamerNonHuman$SeqId, fixed = TRUE)

soma_data_replication_PASS = soma_data_replication_PASS %>% dplyr::select(!AptamerNonHuman[,1])
```

# Merge the SomaScan data with the clinical data
```{r}
#Discovery cohort
merged_data_discovery <- inner_join(soma_data_discovery_PASS, clinical_data_discovery, by = c("SampleId" = "ID"))

#Replication cohort
merged_data_replication <- inner_join(soma_data_replication_PASS, clinical_data_replication, by = c("SampleId" = "ID"))
```

# Identify outliers in the SomaScan data
```{r}
#Discovery cohort
# identify outliers based on mean signal
merged_data_trim <- merged_data_discovery %>% dplyr::select(SampleId, starts_with("seq."))
rownames(merged_data_trim) <- merged_data_trim[,1] #convert sampleIDs into row names
merged_data_trim <- merged_data_trim %>% 
  dplyr::select(-SampleId) %>% log2() 

mean_signal <- rowMeans(merged_data_trim, na.rm = TRUE)
Q3 <- quantile(mean_signal, 0.75)
IQR_val <- IQR(mean_signal)
upper <- Q3 + 3*IQR_val   

suspects <- names(mean_signal[mean_signal > upper])

# create a boxplot to visualize the outliers
jpeg('discovery_outliers.jpg', res = 150, pointsize = 8)
boxplot(mean_signal,  xlab = "Discovery", ylab="Mean protein level", ylim = range(11.8:12.8), yaxs = "i")
abline(h=upper, col="red", lwd=2, lty="dotdash")
dev.off()

# remove the outliers from the merged data
merged_data_discovery_clean <- merged_data_discovery %>% filter(!SampleId %in% suspects)


#Replication cohort
# identify outliers based on mean signal
merged_data_trim <- merged_data_replication %>% dplyr::select(SampleId, starts_with("seq."))
rownames(merged_data_trim) <- merged_data_trim[,1] #convert sampleIDs into row names
merged_data_trim <- merged_data_trim %>%
  dplyr::select(-SampleId) %>% log2()      

mean_signal <- rowMeans(merged_data_trim, na.rm = TRUE)
Q3 <- quantile(mean_signal, 0.75)
IQR_val <- IQR(mean_signal)
upper <- Q3 + 3*IQR_val   

suspects <- names(mean_signal[mean_signal > upper])

# create a boxplot to visualize the outliers
jpeg('replication_outliers.jpg', res = 150, pointsize = 8)
boxplot(mean_signal,  xlab = "Replication", ylab="Mean protein level", ylim = range(11.8:12.8), yaxs = "i")
abline(h=upper, col="red", lwd=2, lty="dotdash")
dev.off()

# remove the outliers from the merged data
merged_data_replication_clean <- merged_data_replication %>% filter(!SampleId %in% suspects)
```

# Calculate surrogate variables for the SomaScan data
```{r}
#Discovery cohort
# prepare the data for surrogate variable analysis
merged_data_log <- merged_data_discovery_clean %>% dplyr::select(SampleId, starts_with("seq."))
rownames(merged_data_log) <- merged_data_log[,1] #convert sampleIDs into row names
merged_data_log <- merged_data_log %>%
  dplyr::select(-SampleId) %>% log2()

# create a model matrix for the clinical variables
clinical_vars <- merged_data_discovery_clean %>% dplyr::select(SampleId, event, PlateId)
rownames(clinical_vars) <- clinical_vars[,1] 

# perform surrogate variable analysis
X <- t(as.matrix(merged_data_log))
mod  <- model.matrix(~ as.factor(event), data = clinical_vars)
mod0 <- model.matrix(~ 1, data = clinical_vars)

svobj <- sva(
  X,
  mod,
  mod0,
  n.sv = 3
)

# Add the surrogate variables to the clinical data
merged_data_discovery_clean$SV1 <- svobj$sv[,1]
merged_data_discovery_clean$SV2 <- svobj$sv[,2]
merged_data_discovery_clean$SV3 <- svobj$sv[,3]

#Replication cohort
# prepare the data for surrogate variable analysis
merged_data_log <- merged_data_replication_clean %>% dplyr::select(SampleId, starts_with("seq."))
rownames(merged_data_log) <- merged_data_log[,1] #convert sampleIDs into row names
merged_data_log <- merged_data_log %>%
  dplyr::select(-SampleId) %>% log2()

# create a model matrix for the clinical variables
clinical_vars <- merged_data_replication_clean %>% dplyr::select(SampleId, event, PlateId)
rownames(clinical_vars) <- clinical_vars[,1] 

# perform surrogate variable analysis
X <- t(as.matrix(merged_data_log))
mod  <- model.matrix(~ as.factor(event), data = clinical_vars)
mod0 <- model.matrix(~ 1, data = clinical_vars)

svobj <- sva(
  X,
  mod,
  mod0,
  n.sv = 3
)

# Add the surrogate variables to the clinical data
merged_data_replication_clean$SV1 <- svobj$sv[,1]
merged_data_replication_clean$SV2 <- svobj$sv[,2]
merged_data_replication_clean$SV3 <- svobj$sv[,3]
```

# Z- and log2 transform the SomaScan data 
```{r}
# function to identify SomaScan protein columns and to center and scale the data
is_seq <- function(.x) grepl("^seq\\.[0-9]{4}", .x) 
cs <- function(.x) {    
  out <- .x - mean(.x)  
  out / sd(out)        
}

#Discovery cohort
# prepare final data set for analysis
cleanData_discovery <- merged_data_discovery_clean %>%
  drop_na(event) %>%                            # rm NAs if present
  mutate(across(matches("^seq\\.[0-9]{4}"), ~ cs(log2(.x))))


#Replication cohort
# prepare final data set for analysis
cleanData_replication <- merged_data_replication_clean %>%
  drop_na(event) %>%                            # rm NAs if present
  mutate(across(matches("^seq\\.[0-9]{4}"), ~ cs(log2(.x))))    
```

# Survival analysis Discovery cohort
```{r}
# keep only columns of interest
cleanData_survival <- cleanData_discovery %>% 
  dplyr::select(SampleId,event,time.event,sex,age_baseline,SV1,SV2,SV3,starts_with("seq."))

cleanData_survival$event <- with(cleanData_survival, ifelse(event == "normal", 0, 
                                                      ifelse(event == "dementia", 1, NA)))

TABLE <- as.data.frame(cleanData_survival)
TABLE$time.event <- as.numeric(TABLE$time.event)
TABLE <- drop_na(TABLE,event)
	
	coefficients<-as.data.frame(matrix(ncol= 10))
 
	names(coefficients) <- c("AptName","Coeff", "se", "Pvalue", "Cox.zphPVal", "N", "ov.lik.ratio","logrank", "r2", "BIC" )
 
	for (i in 9:ncol(TABLE)) {
	print(colnames(TABLE)[i])
  	protein <- TABLE[,c(i,1:8)]
	j= i-8

# test Cox model - if there is an error then put NoConverge into results
  	testCox = try(coxph(Surv(protein$time.event, protein$event) ~ protein[,1] + protein$age_baseline + protein$sex + protein$SV1 + protein$SV2 + protein$SV3, data=protein), 
  	              silent = T)
  	
  	if(class(testCox)[1]=="try-error"){
  	  coefficients[j,1]<- paste(colnames(TABLE)[i])
  	  coefficients[j,2:9] <- "NoConverge"
  	} else {
  	
  	# if Cox model is not an error, run and get the coefficients
  	model.cox<- coxph(Surv(protein$time.event, protein$event) ~ protein[,1] + protein$age_baseline + protein$sex + protein$SV1 + protein$SV2 + protein$SV3, data=protein)
  	
  	testkmz <- try(kmz<- cox.zph(model.cox, transform = "km"), silent = T)
  	
  	coefficients[j,1]<- paste(colnames(TABLE)[i])
  	coefficients[j,2]<- summary(model.cox)$coefficients[1,1]
  	coefficients[j,3]<- summary(model.cox)$coefficients[1,3]
  	coefficients[j,4]<- summary(model.cox)$coefficients[1,5]
  	coefficients[j,6]<- model.cox$n
  	coefficients[j,7]<- summary(model.cox)$logtest[[1]]
  	coefficients[j,8]<- summary(model.cox)$sctest[[1]]
  	coefficients[j,9]<- summary(model.cox)$rsq[[1]] # nagelkerke r square
  	coefficients[j,10] <- BIC(model.cox)
  	
  	# if kmz is not an error, get the Cox.zphPVal
  	if(class(testkmz)[1] == "try-error"){
  	coefficients[j,5]<- "solveError"
  	} else {
  	coefficients[j,5]<- kmz$table[1,3]
  	}
  	}
  	}

# Add protein names to the coefficients table
    prot_list <- getAnalyteInfo(soma_data_discovery_PASS)%>% dplyr::select(AptName, EntrezGeneSymbol)
	coefficients$Protein <- prot_list$EntrezGeneSymbol[match(coefficients$AptName,prot_list$AptName)]
	  
# FDR adjusted P-values
	coefficients$AdjP <- p.adjust(coefficients$Pvalue, method = "fdr")

# Save output for meta-analysis
    fwrite(coefficients,"discovery_SomaScan_survival_SVs.txt", sep = "\t", quote = F, row.names = F, col.names = T) #full output

        #nominally significant protein list for meta-analysis
        sig_proteins_cph_dementia <- coefficients %>% filter(Pvalue<0.05)
        sig_proteins_cph_dementia <- sig_proteins_cph_dementia$AptName

        write.table(sig_proteins_cph_dementia, "discovery_nominal_associations_dementia_cph_SVs.txt", quote = F, row.names = F, col.names = F) 
```

# Survival analysis Replication cohort
```{r}
#Select proteins significant in the Discovery cohort
cleanData_survival <- cleanData_replication %>% 
  dplyr::select(SampleId,event,time.event,sex,age_baseline,SV1,SV2,SV3,all_of(sig_proteins_cph_dementia))     

cleanData_survival$event <- with(cleanData_survival, ifelse(event == "normal", 0, 
                                                      ifelse(event == "dementia", 1, NA)))

TABLE <- as.data.frame(cleanData_survival)
TABLE$time.event <- as.numeric(TABLE$time.event)
TABLE <- drop_na(TABLE,event)
	
	coefficients<-as.data.frame(matrix(ncol= 10))
 
	names(coefficients) <- c("AptName","Coeff", "se", "Pvalue", "Cox.zphPVal", "N", "ov.lik.ratio","logrank", "r2", "BIC" )
 
	for (i in 9:ncol(TABLE)) {
	print(colnames(TABLE)[i])
  	protein <- TABLE[,c(i,1:8)]
	j= i-8

# test Cox model - if there is an error then put NoConverge into results
  	testCox = try(coxph(Surv(protein$time.event, protein$event) ~ protein[,1] + protein$age_baseline + protein$sex + protein$SV1 + protein$SV2 + protein$SV3, data=protein), 
  	              silent = T)
  	
  	if(class(testCox)[1]=="try-error"){
  	  coefficients[j,1]<- paste(colnames(TABLE)[i])
  	  coefficients[j,2:9] <- "NoConverge"
  	} else {
  	
  	# if Cox model is not an error, run and get the coefficients
  	model.cox<- coxph(Surv(protein$time.event, protein$event) ~ protein[,1] + protein$age_baseline + protein$sex + protein$SV1 + protein$SV2 + protein$SV3, data=protein)
  	
  	testkmz <- try(kmz<- cox.zph(model.cox, transform = "km"), silent = T)
  	
  	coefficients[j,1]<- paste(colnames(TABLE)[i])
  	coefficients[j,2]<- summary(model.cox)$coefficients[1,1]
  	coefficients[j,3]<- summary(model.cox)$coefficients[1,3]
  	coefficients[j,4]<- summary(model.cox)$coefficients[1,5]
  	coefficients[j,6]<- model.cox$n
  	coefficients[j,7]<- summary(model.cox)$logtest[[1]]
  	coefficients[j,8]<- summary(model.cox)$sctest[[1]]
  	coefficients[j,9]<- summary(model.cox)$rsq[[1]] # nagelkerke r square
  	coefficients[j,10] <- BIC(model.cox)
  	
  	# if kmz is not an error, get the Cox.zphPVal
  	if(class(testkmz)[1] == "try-error"){
  	coefficients[j,5]<- "solveError"
  	} else {
  	coefficients[j,5]<- kmz$table[1,3]
  	}
  	}
  	}

# Add protein names to the coefficients table
    prot_list <- getAnalyteInfo(soma_data_replication_PASS) %>% dplyr::select(AptName, EntrezGeneSymbol)
	coefficients$Protein <- prot_list$EntrezGeneSymbol[match(coefficients$AptName,prot_list$AptName)]
	  
# FDR adjusted P-values
	coefficients$AdjP <- p.adjust(coefficients$Pvalue, method = "fdr")

# Save output for meta-analysis
    fwrite(coefficients,"replication_SomaScan_survival_SVs.txt", sep = "\t", quote = F, row.names = F, col.names = T) #full output
```

# Meta-analysis of the discovery and replication cohorts
```{r}
# load the discovery and replication results
discovery_results <- fread("discovery_SomaScan_survival_SVs.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
replication_results <- fread("replication_SomaScan_survival_SVs.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)  

discovery_results <- discovery_results %>% dplyr::select(AptName,Coeff,se,Pvalue,N,logrank)
replication_results <- replication_results %>% dplyr::select(AptName,Coeff,se,Pvalue,N,logrank)

# keep only nominally significant from Discovery cohort (=replication)
discovery_results <- semi_join(discovery_results, replication_results, by = "AptName")

# merge summary statistics on AptName
discovery_results$Study <- "TPD"
replication_results$Study <- "OPDC"

combined_data_dementia <- rbind(discovery_results, replication_results)
combined_data_dementia$Study <- as.factor(combined_data_dementia$Study)
combined_data_dementia <- combined_data_dementia %>% drop_na()

# meta-analysis
meta_results <- list()

for(protein in unique(combined_data_dementia$AptName)) {
  
  # Subset data for the current protein
  protein_data <- combined_data_dementia[combined_data_dementia$AptName == protein, ]
  
  # Run meta-analysis for the current protein
  res <- rma(yi = protein_data$Coeff, sei = protein_data$se, method = "REML", slab = protein_data$Study, data = protein_data)
  
  # Store the result in the list
  meta_results[[protein]] <- res
}

# Create a data frame to store the meta-analysis results for each protein
meta_analysis_summary <- data.frame(
  AptName = character(),
  logHR = numeric(),
  SE = numeric(),
  CI_low = numeric(),
  CI_high = numeric(),
  pval = numeric(),
  I2 = numeric(),
  Tau2 = numeric(),
  Q = numeric(),
  Q_pval = numeric(),  
  stringsAsFactors = FALSE
)

# Loop through the list and extract relevant information for each protein
for(protein in names(meta_results)) {
  
  # Extract the result for the current protein
  res <- meta_results[[protein]]
  
  # Store the summary results in the data frame
  meta_analysis_summary <- rbind(meta_analysis_summary, data.frame(
    AptName = protein,
    logHR = res$b,  # Pooled logHR
    SE = res$se,    # Standard error
    CI_low = res$ci.lb,  # Lower bound of CI
    CI_high = res$ci.ub, # Upper bound of CI
    pval = res$pval, # p-value
    I2 = res$I2,	           # Heterogeneity I²
    Tau2 = res$tau2,	       # Between-study variance
    Q = res$QE,	            # Q-statistic
    Q_pval = res$QEp       # Q-test p-value	
  ))
}

# Add gene symbols to the meta-analysis summary
meta_analysis_summary$geneSymbol <- prot_list$EntrezGeneSymbol[match(meta_analysis_summary$AptName,prot_list$AptName)]

# Adjust p-values for multiple testing using the Benjamini-Hochberg method
meta_analysis_summary <- meta_analysis_summary %>%
  mutate(Padj = p.adjust(pval, method = "BH"))

# Save the meta-analysis summary results to a file
write.table(meta_analysis_summary,"meta_pdd_cph.txt", sep = "\t", quote = F, row.names = F, col.names = T)

```

# Plot volcano plot for meta-analysis results
```{r}
# Create a volcano plot for the meta-analysis results
pval_threshold <- 0.05

meta_analysis_summary$color_group <- "grey"  # Default: non-significant
meta_analysis_summary$color_group[meta_analysis_summary$Padj < pval_threshold] <- "red"
meta_analysis_summary$color_group[meta_analysis_summary$Padj > pval_threshold & meta_analysis_summary$pval < pval_threshold] <- "grey"

volcano_plot_survival_meta <- ggplot(meta_analysis_summary, aes(x = exp(logHR), y = -log10(Padj), label = geneSymbol, color = color_group)) +
  geom_point(alpha = 0.8, size = 1.3) +
  scale_color_manual(values = c("grey" = "grey", "red" = "red", "grey" = "grey")) +
  geom_hline(yintercept = -log10(pval_threshold), linetype = "dashed", color = "grey") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey") +
  geom_text_repel(data = subset(meta_analysis_summary, Padj < pval_threshold), size = 3, max.overlaps = 15, color = "black", force = 2, force_pull = 0.5, min.segment.length = 0.2) +
  labs(x = "Hazard Ratio", y = "-log10(FDR)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") + 
  theme(aspect.ratio = 1,
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"),
        legend.position = "none",  # Show legend
        legend.title = element_blank())

ggsave("volcano_plot_survival_meta.png", 
       volcano_plot_survival_meta, 
       width = 5, height = 5, dpi = 300)
```

# Corres plot of the meta-analysis results
```{r}
# Step 1: Find common row names
common_rows <- intersect(discovery_results$AptName, replication_results$AptName)

# Step 2: Subset both data frames to common rows
df1_common <- discovery_results %>% filter(AptName %in% common_rows)
df2_common <- replication_results %>% filter(AptName %in% common_rows)

df1_common <- df1_common[order(df1_common$AptName), ]
df2_common <- df2_common[order(df2_common$AptName), ]
all(df1_common$AptName == df2_common$AptName)

# Step 3: Merge data
merged <- merge(df1_common, df2_common, by = "AptName", suffixes = c("_disc", "_repl"))
names(merged)[1] <- "Protein"
merged$Gene <- prot_list$EntrezGeneSymbol[match(merged$Protein,prot_list$AptName)]
merged$FDR <- meta_analysis_summary$Padj[match(merged$Protein,meta_analysis_summary$AptName)]

# Step 4: Identify significant genes
merged <- merged %>% 
   mutate(significant=ifelse(FDR < 0.05, "yes", "no"))

# Step 5: Calculate z-score of the coefficients in discovery and replication cohorts
merged$z_disc <- merged$Coeff_disc / merged$se_disc
merged$z_repl <- merged$Coeff_repl / merged$se_repl

# Step 6: Create the corres plot
significant_proteins <- subset(merged, significant == "yes")

corres_plot_z <- ggplot(merged, aes(x = z_disc, y = z_repl)) +
 geom_point(aes(color = significant), size = 1.5, alpha = 0.8) +
 geom_smooth(data = significant_proteins, method = "lm", color = "blue", se = TRUE, fill = "gray") +
 scale_color_manual(values = c("gray", "red")) +
 geom_point(data=subset(merged, significant=="yes"), color="red", size=1.3) + 
 geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.2) +
 geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.2) +
 geom_text_repel(data = subset(merged, significant == "yes"), aes(label = Gene), size = 3, max.overlaps = 15, force_pull = 0.5, force = 2, min.segment.length = 0.2) +
 labs(x = "z-score (Discovery)",
 y = "z-score (Replication)") +
 theme_minimal(base_size = 14) +
 guides(color = "none") +
 theme(aspect.ratio = 1) +
 theme(legend.position = "none") +
 theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
 panel.background = element_blank(), axis.line = element_line(colour = "black"))

ggsave("corres_plot_z.png", 
       corres_plot_z, 
       width = 5, height = 5, dpi = 300)

# Step 7: Calculate correlation between effect size of significant proteins in the meta-analysis
cor(significant_proteins$z_disc, significant_proteins$z_repl)
cor.test(significant_proteins$z_disc, significant_proteins$z_repl)
```

# Differential abundance analysis of the SomaScan data by APOE status in the discovery cohort
```{r}
# Step 1: Create a matrix of the expression data

cleanData_discovery_exprs <- cleanData_discovery %>% dplyr::select(SampleId, starts_with("seq."))
cleanData_discovery_exprs <- as.data.frame(cleanData_discovery_exprs) 
rownames(cleanData_discovery_exprs) <- NULL

cleanData_discovery_exprs_t <- setNames(data.frame(t(cleanData_discovery_exprs[ , - 1])), cleanData_discovery_exprs[ , 1])
cleanData_discovery_exprs_t <- cleanData_discovery_exprs_t %>% dplyr::select(order(colnames(cleanData_discovery_exprs_t))) #order columns alphabetically

write.table(cleanData_discovery_exprs_t,"discovery_exprsData.txt", sep = "\t", quote = F, row.names = T, col.names = T)

exprsFile <- "discovery_exprsData.txt"
exprs_discovery <- as.matrix(read.table(exprsFile, header=TRUE, sep="\t",
                               row.names=1,
                               as.is=TRUE))

# Step 2: Create a table for the phenotype data
pheno_discovery <- cleanData_discovery %>% dplyr::select(SampleId, age_baseline, sex, E4_allele, SV1, SV2, SV3)
pheno_discovery$E4_allele <- with(pheno_discovery, ifelse(E4_allele == "e4-carrier", 2,1))   

write.table(pheno_discovery,"discovery_pData.txt", sep = "\t", quote = F, row.names = F, col.names = T)

pDataFile <- "discovery_pData.txt"
pData_discovery <- read.table(pDataFile,
                     row.names=1, header=TRUE, sep="\t")

pData_discovery <- pData_discovery[colnames(exprs_discovery), , drop = FALSE] #reorder pData IDs to match exprs
all(rownames(pData_discovery)==colnames(exprs_discovery)) #This must be true

# Step 3: Assemble ExpressionSet object
phenoData_discovery <- new("AnnotatedDataFrame", data = pData_discovery)

dataSet_discovery <- ExpressionSet(assayData=exprs_discovery,
                   phenoData=phenoData_discovery)

dataSet_discovery <- dataSet_discovery[, !is.na(pData_discovery$E4_allele)]  # Remove rows with NA in E4_allele
dataSet_discovery$E4_allele = factor(dataSet_discovery$E4_allele)

# Step 4: Create a design matrix for the linear model
design.cov <- model.matrix(~ 0 + E4_allele + sex + age_baseline + SV1 + SV2 + SV3, data=dataSet_discovery)
head(design.cov)
colnames(design.cov) <- c("E4_negative","E4_positive","sex","age","sv1","sv2","sv3")

fit.cov <- lmFit(dataSet_discovery, design.cov)
contrast_cov <- makeContrasts( 
                     E4posvE4neg = E4_positive - E4_negative,
                     levels = design.cov)
fitC.cov <- contrasts.fit(fit.cov, contrast_cov)
fitC.cov <- eBayes(fitC.cov) #calculate t-statistics

saveRDS(fitC.cov,"discovery_Lm_APOE_cov.rds")

results.cov_discovery <- decideTests(fitC.cov) #summarize results

volcano_cov_tbl <- topTable(fitC.cov, sort = "none", n = Inf)
volcano_cov_tbl <- cbind(geneSymbol = prot_list$EntrezGeneSymbol[ match(rownames(volcano_cov_tbl), prot_list$AptName) ],
      volcano_cov_tbl)

write.table(volcano_cov_tbl,"discovery_Lm_APOE_cov.txt", row.names = T, col.names = T, sep ="\t", quote = F)

sig_proteins_lm_apoe <- volcano_cov_tbl %>% filter(P.Value<0.05)
sig_proteins_lm_apoe <- row.names(sig_proteins_lm_apoe)

write.table(sig_proteins_lm_apoe, "discovery_nominal_associations_apoe_lm.txt", quote = F, row.names = F, col.names = F)
```

# Differential protein abundance analysis of the SomaScan data by APOE status in the replication cohort
```{r}
nom_p_apoe_discovery <- fread("discovery_nominal_associations_apoe_lm.txt", header = F)

# Step 1: Create a matrix of the expression data        
cleanData_replication_exprs <- cleanData_replication %>% dplyr::select(SampleId, all_of(nom_p_apoe_discovery$V1))
cleanData_replication_exprs <- as.data.frame(cleanData_replication_exprs)
rownames(cleanData_replication_exprs) <- NULL

cleanData_replication_exprs_t <- setNames(data.frame(t(cleanData_replication_exprs[ , - 1])), cleanData_replication_exprs[ , 1])
cleanData_replication_exprs_t <- cleanData_replication_exprs_t %>% dplyr::select(order(colnames(cleanData_replication_exprs_t))) #order columns alphabetically              

write.table(cleanData_replication_exprs_t,"replication_exprsData.txt", sep = "\t", quote = F, row.names = T, col.names = T)

exprsFile <- "replication_exprsData.txt"
exprs_replication <- as.matrix(read.table(exprsFile, header=TRUE, sep="\t",
                               row.names=1,
                               as.is=TRUE))

# Step 2: Create a table for the phenotype data
pheno_replication <- cleanData_replication %>% dplyr::select(SampleId, age_baseline, sex, E4_allele, SV1, SV2, SV3)
pheno_replication$E4_allele <- with(pheno_replication, ifelse(E4_allele == "e4-carrier", 2,1))   

pheno_replication <- pheno_replication %>% arrange(SampleId)
pheno_replication <- pheno_replication %>% drop_na(E4_allele)

write.table(pheno_replication,"replication_pData.txt", sep = "\t", quote = F, row.names = F, col.names = T)

pDataFile <- "replication_pData.txt"
pData_replication <- read.table(pDataFile,
                     row.names=1, header=TRUE, sep="\t")

common_samples <- intersect(colnames(exprs_replication), rownames(pData_replication))
exprs_replication <- exprs_replication[, common_samples] #keep only common samples 
pData_replication <- pData_replication[colnames(exprs_replication), , drop = FALSE] #reorder pData IDs to match exprs
all(rownames(pData_replication)==colnames(exprs_replication)) #This must be true

# Step 3: Assemble ExpressionSet object
phenoData_replication <- new("AnnotatedDataFrame", data = pData_replication)

dataSet_replication <- ExpressionSet(assayData=exprs_replication,
                   phenoData=phenoData_replication)

dataSet_replication <- dataSet_replication[, !is.na(pData_replication$E4_allele)]  # Remove rows with NA in E4_allele
dataSet_replication$E4_allele = factor(dataSet_replication$E4_allele)

# Step 4: Create a design matrix for the linear model
design.cov <- model.matrix(~ 0 + E4_allele + sex + age_baseline + SV1 + SV2 + SV3, data=dataSet_replication)
head(design.cov)
colnames(design.cov) <- c("E4_negative","E4_positive","sex","age","sv1","sv2","sv3")

fit.cov <- lmFit(dataSet_replication, design.cov)
contrast_cov <- makeContrasts( 
                     E4posvE4neg = E4_positive - E4_negative,
                     levels = design.cov)
fitC.cov <- contrasts.fit(fit.cov, contrast_cov)
fitC.cov <- eBayes(fitC.cov) #calculate t-statistics

saveRDS(fitC.cov,"replication_Lm_APOE_cov.rds")

results.cov_replication <- decideTests(fitC.cov) #summarize results

volcano_cov_tbl <- topTable(fitC.cov, sort = "none", n = Inf)
volcano_cov_tbl <- cbind(geneSymbol = prot_list$EntrezGeneSymbol[ match(rownames(volcano_cov_tbl), prot_list$AptName) ],
      volcano_cov_tbl)

write.table(volcano_cov_tbl, "replication_Lm_APOE_cov.txt", row.names = T, col.names = T, sep = "\t", quote = F)

sig_proteins_lm_apoe <- volcano_cov_tbl %>% filter(P.Value<0.05)
sig_proteins_lm_apoe <- row.names(sig_proteins_lm_apoe)

write.table(sig_proteins_lm_apoe, "replication_nominal_associations_apoe_lm.txt", quote = F, row.names = F, col.names = F)
```

# Meta-analysis of the discovery and replication cohorts for differential protein abundance analysis by APOE status
```{r}
# load the discovery and replication results
discovery_results <- readRDS("discovery_Lm_APOE_cov.rds")
replication_results <- readRDS("replication_Lm_APOE_cov.rds")

# Discovery cohort
## Calculate standard errors for the discovery study
SE_discovery <- sqrt(discovery_results$s2.post) * discovery_results$stdev.unscaled
## Create a data frame for the discovery study
meta_data_discovery <- data.frame(
  protein = rownames(discovery_results$coefficients), # Effect sizes
  logFC = as.vector(discovery_results$coefficients),  # Standard errors
  SE = as.vector(SE_discovery),
  Study = "discovery"
)

# Replication cohort
## Calculate standard errors for the replication study
SE_replication <- sqrt(replication_results$s2.post) * replication_results$stdev.unscaled
## Create a data frame for the replication study
meta_data_replication <- data.frame(
  protein = rownames(replication_results$coefficients),
  logFC = as.vector(replication_results$coefficients),  
  SE = as.vector(SE_replication),  
  Study = "replication"
)

# Combine the discovery and replication data frames
combined_meta_data_apoe <- rbind(meta_data_discovery, meta_data_replication)

# Run meta-analysis for each protein
meta_results_list <- list()

for (protein in unique(combined_meta_data_apoe$protein)) {
  protein_data <- combined_meta_data_apoe[combined_meta_data_apoe$protein == protein, ]

  # Run meta-analysis for the current protein
if (nrow(protein_data) > 1) {
  meta_results <- rma(yi = logFC, sei = SE, 
                    data = protein_data, 
                    method = "REML")
  meta_results_list[[protein]] <- meta_results
} else {
    meta_results_list[[protein]] <- NA  # Not enough data for meta-analysis
  }
}

# Create a summary data frame for the meta-analysis results
meta_summary_apoe <- data.frame()

for (protein in names(meta_results_list)) {
  meta_result <- meta_results_list[[protein]]
  
  # Ensure meta_result is a valid rma object before proceeding
  if (inherits(meta_result, "rma")) {
    summary_stats <- data.frame(
      Protein = protein,
      Estimate = coef(meta_result),  # Effect size (logFC)
      SE = meta_result$se,           # Standard Error
      CI_lower = meta_result$ci.lb,  # 95% CI Lower Bound
      CI_upper = meta_result$ci.ub,  # 95% CI Upper Bound
      Pvalue = meta_result$pval,     # P-value
      I2 = meta_result$I2,           # Heterogeneity I²
      Tau2 = meta_result$tau2,       # Between-study variance
      Q = meta_result$QE,            # Q-statistic
      Q_pval = meta_result$QEp       # Q-test p-value
    )
    
    meta_summary_apoe <- rbind(meta_summary_apoe, summary_stats)
  }
}

# Adjust P-values for multiple testing (FDR)
meta_summary_apoe$FDR <- p.adjust(meta_summary_apoe$Pvalue, method = "BH")

meta_summary_apoe$Gene <- prot_list$EntrezGeneSymbol[match(meta_summary_apoe$Protein,prot_list$AptName)]

write.table(meta_summary_apoe, "meta_APOE_Lm.txt", row.names = F, col.names = T, sep = "\t", quote = F)

```

# Plot volcano plot for meta-analysis results of differential abundance analysis by APOE status
```{r}
# Create a volcano plot for the meta-analysis results of differential abundance analysis by APOE status
# Define significance thresholds
logFC_threshold <- 0.5
pval_threshold <- 0.05

meta_summary_apoe$color_group <- "grey"  # Default: non-significant
meta_summary_apoe$color_group[meta_summary_apoe$FDR < pval_threshold & abs(meta_summary_apoe$Estimate) > logFC_threshold] <- "red"
meta_summary_apoe$color_group[meta_summary_apoe$FDR < pval_threshold & abs(meta_summary_apoe$Estimate) <= logFC_threshold] <- "green4"


volcano_plot_Lm_apoe_meta <- ggplot(meta_summary_apoe, aes(x = Estimate, y = -log10(FDR), label = Gene, color = color_group)) +
  geom_point(alpha = 0.8, size = 1.5) +
  scale_color_manual(values = c("grey" = "grey", "red" = "red", "green4" = "green4")) +
  geom_hline(yintercept = -log10(pval_threshold), linetype = "dashed", color = "grey") +
  geom_vline(xintercept = c(-logFC_threshold, logFC_threshold), linetype = "dashed", color = "grey") +
  geom_text_repel(data = subset(meta_summary_apoe, FDR < pval_threshold), size = 4, max.overlaps = 30, color = "black") +
  labs(x = "LogFC", y = "-log10(FDR)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") + 
  theme(aspect.ratio = 1,
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"),
        legend.position = "none",  # Show legend
        legend.title = element_blank())

ggsave("volcano_plot_Lm_apoe_meta.png", 
       volcano_plot_Lm_apoe_meta, 
       width = 5, height = 5, dpi = 300)
```

# Plot corres plot for meta-analysis results of differential abundance analysis by APOE status
```{r}
# Load the discovery and replication results
discovery_results <- readRDS("discovery_Lm_APOE_cov.rds")
replication_results <- readRDS("replication_Lm_APOE_cov.rds")

# Step 1: Find common row names
discovery_tbl <- topTable(discovery_results,sort="none",n=Inf)
replication_tbl <- topTable(replication_results,sort="none",n=Inf)

common_rows <- intersect(rownames(discovery_tbl), rownames(replication_tbl))

# Step 2: Subset both data frames to common rows
df1_common <- discovery_tbl[common_rows, , drop = FALSE]
df2_common <- replication_tbl[common_rows, , drop = FALSE]

# Step 3: Merge data
df1_common$AptName <- rownames(df1_common)
df2_common$AptName <- rownames(df2_common)

df1_common <- df1_common[order(df1_common$AptName), ]
df2_common <- df2_common[order(df2_common$AptName), ]
all(df1_common$AptName == df2_common$AptName)

merged <- merge(df1_common, df2_common, by = "AptName", suffixes = c("_disc", "_repl"))
merged$FDR <- meta_summary_apoe$FDR[match(merged$AptName,meta_summary_apoe$Protein)]
merged$Pvalue <- meta_summary_apoe$Pvalue[match(merged$AptName,meta_summary_apoe$Protein)]

merged$Gene <- prot_list$EntrezGeneSymbol[match(merged$AptName,prot_list$AptName)]

# Step 4: Identify significant genes
merged <- merged %>% 
   mutate(significant=ifelse(FDR < 0.05, "yes", 
                            ifelse(Pvalue <0.05, "nominally", "no")))

merged$significant <- factor(merged$significant, levels = c("no", "nominally", "yes"))

significant_proteins <- subset(merged, significant == "yes")

# Step 5: Create the corres plot
corres_plot_apoe <- ggplot(merged, aes(x = logFC_disc, y = logFC_repl)) +
 geom_point(aes(color = significant), size = 1.5, alpha = 0.8) +
 geom_smooth(data = significant_proteins, method = "lm", color = "blue", se = TRUE, fill = "gray") +
 geom_text_repel(data = subset(merged, significant == "yes"), aes(label = Gene), size = 4, max.overlaps = 30, force_pull = 0.5, min.segment.length = 0.2) +
 scale_color_manual(values = c("gray", "grey", "red")) +
 geom_point(data=subset(merged, significant=="yes"), color="red", size=1.3) +
 geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.2) +
 geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.2) +
 labs(x = "logFC (Discovery)",
 y = "logFC (Replication)") +
 theme_minimal(base_size = 14) +
 guides(color = "none") +
 theme(aspect.ratio = 1) +
 theme(legend.position = "none") +
 theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
 panel.background = element_blank(), axis.line = element_line(colour = "black"))

ggsave("corres_plot_apoe.png", 
       corres_plot_apoe, 
       width = 5, height = 5, dpi = 300)


# Step 6: Correlation between effect size of significant proteins in the meta-analysis
cor(significant_proteins$logFC_disc, significant_proteins$logFC_repl)
cor.test(significant_proteins$logFC_disc, significant_proteins$logFC_repl)

```

# Plot corres plot of effect sizes in survival meta-analysis CPH vs Lm APOE meta-analysis, FDR<0.05
```{r}
# Load the meta-analysis results
meta_cph_dementia <- fread("meta_pdd_cph.txt")
meta_cph_dementia <- meta_cph_dementia %>% dplyr::select(AptName,logHR,SE,pval,Padj)

meta_apoe4 <- fread("meta_APOE_Lm.txt")
meta_apoe4 <- meta_apoe4 %>% dplyr::select(Protein,Estimate,SE,Pvalue,FDR)
meta_apoe4 <- meta_apoe4 %>% dplyr::rename(AptName = Protein)
                        
# Step 1: Find common row names
common_rows <- intersect(meta_cph_dementia$AptName, meta_apoe4$AptName)

# Step 2: Subset both data frames to common rows
df1_common <- meta_cph_dementia %>% filter(AptName %in% common_rows)
df2_common <- meta_apoe4 %>% filter(AptName %in% common_rows)

df1_common <- df1_common %>% dplyr::rename(Estimate = logHR,
                                    Pvalue = pval,
                                    FDR = Padj)

# Merge data
merged <- merge(df1_common, df2_common, by = "AptName", suffixes = c("_dementia", "_apoe"))
merged$Gene <- prot_list$EntrezGeneSymbol[match(merged$AptName,prot_list$AptName)]

# Identify significant genes
merged <- merged %>% 
   mutate(significant=ifelse(FDR_apoe < 0.05 & FDR_dementia < 0.05, "yes_both", 
                            ifelse(FDR_apoe < 0.05 & FDR_dementia >= 0.05 & Pvalue_dementia > 0.05, "yes_apoe",
                                ifelse(FDR_dementia < 0.05 & FDR_apoe >= 0.05, "yes_dementia", 
                                       ifelse(FDR_apoe < 0.05 & Pvalue_dementia <0.05 & FDR_dementia > 0.05, "yes_apoe_nominal", "no")))))

merged$z_dementia <- merged$Estimate_dementia / merged$SE_dementia
merged$z_apoe <- merged$Estimate_apoe / merged$SE_apoe

corres_plot_apoe_dementia_z <- ggplot(merged, aes(x = z_dementia, y = z_apoe, color = significant)) +
 scale_color_manual(name='Legend',
                    breaks=c('N.S.', 'FDR_dementia <0.05 & FDR_apoe <0.05', 'FDR_apoe <0.05', 'FDR_dementia <0.05', 'FDR_apoe <0.05 & Pval_dementia <0.05'),
                    values = c("N.S." = "gray", "FDR_dementia <0.05 & FDR_apoe <0.05" = "red", "FDR_apoe <0.05" = "green3", "FDR_dementia <0.05" = "orange", "FDR_apoe <0.05 & Pval_dementia <0.05" = "blue3")) +
 geom_point(data=subset(merged, significant=="yes_both"), color="red", size=1.3) +
  geom_point(data=subset(merged, significant=="yes_apoe"), color="green3", size=1.3, alpha = 0.8) +
  geom_point(data=subset(merged, significant=="yes_dementia"), color="orange", size=1.3, alpha = 0.8) +
  geom_point(data=subset(merged, significant=="yes_apoe_nominal"), color="blue3", size=1.3, alpha = 0.8) +
  geom_text_repel(data = subset(merged, significant == "yes_both"), aes(label = Gene), size = 4, max.overlaps = 20, force_pull = 0.2, force = 2, min.segment.length = 0.1, nudge_x = 0.5, color = "black") +
  geom_text_repel(data = subset(merged, significant == "yes_apoe_nominal"), aes(label = Gene), size = 4, max.overlaps = 20, force_pull = 0.2, force = 2, min.segment.length = 0.1, nudge_x = 0.5, color = "black") +
 geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.2) +
 geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.2) +
 labs(x = "z-score of effect size (survival)",
 y = "z-score of effect size (APOE4)") +
 theme_minimal(base_size = 14) +
 theme(aspect.ratio = 1) +
 theme(legend.position = "right") +
 theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
 panel.background = element_blank(), axis.line = element_line(colour = "black"))

ggsave("corres_plot_apoe_dementia_z.png", 
       corres_plot_apoe_dementia_z, 
       width = 5, height = 5, dpi = 300)
```

# pQTL analysis of TBCA and S100A13 in the discovery cohort
```{r}
# Upload principal components for the discovery cohort
pca_discovery <- fread("discovery_PCA.eigenvec", header = FALSE)

# Prepare the data for pQTL analysis
cleanData_survival <- cleanData_discovery %>%
  dplyr::select(SampleId, sex, age_baseline, rs429358, starts_with("seq."))
  
TABLE <- as.data.frame(cleanData_survival)
    TABLE <- TABLE %>% drop_na(rs429358)

protein_labels <- c("seq.12501.10" = "TBCA", "seq.7223.60" = "S100A13")

TABLE_long <- TABLE %>% 
  pivot_longer(cols = c(`seq.12501.10`, `seq.7223.60`), names_to = "Protein", values_to = "Value")

TABLE_long$PC1 <- pca_discovery$V3[match(TABLE_long$SampleId, pca_discovery$V2)]
TABLE_long$PC2 <- pca_discovery$V4[match(TABLE_long$SampleId, pca_discovery$V2)]

TABLE_long <- TABLE_long %>%
  mutate(rs429358 = as.numeric(as.character(rs429358)))
TABLE_long$sex <- with(TABLE_long, ifelse(sex == "Male", 1,
                                          ifelse(sex == "Female", 2, NA)))

# Perform linear regression for each protein
pqtl_results <- TABLE_long %>%
  group_by(Protein) %>%
  do(tidy(lm(Value ~ rs429358 + sex + age_baseline + PC1 + PC2, data = .))) 

# Adjust p-values for multiple testing using the Benjamini-Hochberg method
pqtl_results$p_adj <- p.adjust(pqtl_results$p.value, method = "fdr")

# Violin plots
pqtl_plot <- ggplot(TABLE_long, aes(x = rs429358, y = Value, fill = rs429358)) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "black") +
  geom_violin(scale = "width", trim = FALSE, color = "black") +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  facet_wrap(~ Protein, labeller = labeller(Protein = protein_labels), scales = "fixed") +
  scale_fill_manual(values = c("0" = "#1b9e77", "1" = "#d95f02", "2" = "#7570b3")) +
  labs(
    x = "rs429358 genotype", 
    y = "log2 RFU (z-transformed)"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid = element_blank(),
    axis.line.x = element_line(),
    axis.line.y = element_line(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_text(size = 9),
    axis.text.y = element_text(size = 9),
    legend.position = "none",
    axis.title.y = element_text(angle = 90, vjust = 0.5, size = 10),
    strip.text = element_text(face = "bold", size = 10)
  )

ggsave("plot_pqtl_tpd.png", 
       pqtl_plot,
       width = 4, height = 4, 
       dpi = 300)
```

# Mediation analysis of TBCA and S100A13 levels on the association between APOE status and dementia in the discovery cohort
```{r}
# Prepare the data for mediation analysis
cleanData_survival <- cleanData_discovery %>%
  dplyr::select(SampleId, sex, age_baseline, E4_allele, time.event, event, SV1, SV2, SV3, "seq.12501.10", "seq.7223.60")
  
cleanData_survival$E4_allele <- with(cleanData_survival, ifelse(E4_allele == "e4-carrier", 2,1)) 

TABLE <- as.data.frame(cleanData_survival)
TABLE <- na.omit(TABLE[, c("time.event", "event", "E4_allele","seq.12501.10", "seq.7223.60", "age_baseline", "sex", "SV1", "SV2", "SV3")])

TABLE$seq.12501.10 <- as.numeric(scale(TABLE$seq.12501.10))
TABLE$seq.7223.60 <- as.numeric(scale(TABLE$seq.7223.60))
TABLE$E4_allele <- as.numeric(TABLE$E4_allele)
TABLE$sex  <- as.factor(TABLE$sex)
TABLE$event <- ifelse(TABLE$event == "normal", 0, 1)

#TBCA mediation analysis
fit <- cmest(
  data = TABLE,
  model = "rb",          # regression-based approach
  outcome = "time.event",
  event = "event",
  exposure = "E4_allele",
  mediator = "seq.12501.10",
  basec = c("age_baseline", "sex"),

  yreg = "coxph",
  mreg = list("linear"),
  mval = list(0),
  EMint = F,

  inference = "bootstrap",
  nboot = 1000
)

summary(fit)


# S100A13 mediation analysis
fit2 <- cmest(
  data = TABLE,
  model = "rb",          # regression-based approach
  outcome = "time.event",
  event = "event",
  exposure = "E4_allele",
  mediator = "seq.7223.60",
  basec = c("age_baseline", "sex"),

  yreg = "coxph",
  mreg = list("linear"),
  mval = list(0),
  EMint = FALSE,

  inference = "bootstrap",
  nboot = 1000
)

summary(fit2)
```     

# Survival analysis of the SomaScan data adjusted for APOE status in the discovery cohort
```{r}
# Prepare the data for survival analysis
cleanData_survival <- cleanData_discovery %>% 
  dplyr::select(SampleId,event,time.event,sex,age_baseline,E4_allele,SV1,SV2,SV3,starts_with("seq."))

cleanData_survival$event <- with(cleanData_survival, ifelse(event == "normal", 0, 
                                                      ifelse(event == "dementia", 1, NA)))

TABLE <- as.data.frame(cleanData_survival)

TABLE <- TABLE[!is.na(TABLE$E4_allele), ]
TABLE <- drop_na(TABLE,event)

TABLE$E4_allele <- factor(TABLE$E4_allele)
TABLE$time.event <- as.numeric(TABLE$time.event)
TABLE$event <- as.numeric(TABLE$event)

# Run Cox proportional hazards model for each protein
	coefficients<-as.data.frame(matrix(ncol= 10))
 
	names(coefficients) <- c("AptName","Coeff", "se", "Pvalue", "Cox.zphPVal", "N", "ov.lik.ratio","logrank", "r2", "BIC" )
 
	for (i in 10:ncol(TABLE)) {
	print(colnames(TABLE)[i])
  	protein <- TABLE[,c(i,1:9)]
	j= i-9

    #Test Cox model - if there is an error then put NoConverge into results
  	testCox = try(coxph(Surv(protein$time.event, protein$event) ~ protein[,1] + protein$E4_allele + protein$age_baseline + protein$sex + protein$SV1 + protein$SV2 + protein$SV3, data=protein), 
  	              silent = T)
  	
  	if(class(testCox)[1]=="try-error"){
  	  coefficients[j,1]<- paste(colnames(TABLE)[i])
  	  coefficients[j,2:9] <- "NoConverge"
  	} else {
  	
  	#If Cox model is not an error, run and get the coefficients
  	model.cox<- coxph(Surv(protein$time.event, protein$event) ~ protein[,1] + protein$E4_allele + protein$age_baseline + protein$sex + protein$SV1 + protein$SV2 + protein$SV3, data=protein)
  	
  	testkmz <- try(kmz<- cox.zph(model.cox, transform = "km"), silent = T)
  	
  	coefficients[j,1]<- paste(colnames(TABLE)[i])
  	coefficients[j,2]<- summary(model.cox)$coefficients[1,1]
  	coefficients[j,3]<- summary(model.cox)$coefficients[1,3]
  	coefficients[j,4]<- summary(model.cox)$coefficients[1,5]
  	coefficients[j,6]<- model.cox$n
  	coefficients[j,7]<- summary(model.cox)$logtest[[1]]
  	coefficients[j,8]<- summary(model.cox)$sctest[[1]]
  	coefficients[j,9]<- summary(model.cox)$rsq[[1]] # nagelkerke r square
  	coefficients[j,10] <- BIC(model.cox)
  	
  	#If kmz is not an error, get the Cox.zphPVal
  	if(class(testkmz)[1] == "try-error"){
  	coefficients[j,5]<- "solveError"
  	} else {
  	coefficients[j,5]<- kmz$table[1,3]
  	}
  	}
  	}
    
# Add gene symbols to the coefficients data frame
    prot_list <- getAnalyteInfo(soma_data_discovery_PASS) %>% dplyr::select(AptName, EntrezGeneSymbol)
	coefficients$Protein <- prot_list$EntrezGeneSymbol[match(coefficients$AptName,prot_list$AptName)]

# Adjust p-values for multiple testing using the Benjamini-Hochberg method 
    coefficients$AdjP <- p.adjust(coefficients$Pvalue, method = "fdr")

# Save the coefficients data frame to a file
    fwrite(coefficients,"discovery_SomaScan_cph_apoe4_all_SVs.txt", quote = F, row.names = F, col.names = T)

    #nominally significant protein list for meta-analysis
    sig_proteins_cph_dementia_apoe4 <- coefficients %>% filter(Pvalue<0.05)
    sig_proteins_cph_dementia_apoe4 <- sig_proteins_cph_dementia_apoe4$AptName

    write.table(sig_proteins_cph_dementia_apoe4, "discovery_nominal_associations_cph_apoe4_SVs.txt", quote = F, row.names = F, col.names = F)
```

# Survival analysis of the SomaScan data adjusted for APOE status in the replication cohort
```{r}    
# Import significant proteins from the discovery cohort
apoe4_nom_p_cph_discovery <- fread("discovery_nominal_associations_cph_apoe4_SVs.txt", header = FALSE)

# Select proteins significant in the discovery cohort for survival analysis in the replication cohort
cleanData_survival <- cleanData_replication %>% 
  dplyr::select(SampleId,event,time.event,sex,age_baseline,SV1,SV2,SV3,E4_allele,all_of(apoe4_nom_p_cph_discovery$V1))

TABLE <- as.data.frame(cleanData_survival)

TABLE$event <- with(TABLE, ifelse(event == "normal",0,
                                  ifelse(event=="dementia",1,NA)))

TABLE <- TABLE[!is.na(TABLE$E4_allele), ]
TABLE <- drop_na(TABLE,event)

TABLE$E4_allele <- factor(TABLE$E4_allele)
TABLE$time.event <- as.numeric(TABLE$time.event)
TABLE$event <- as.numeric(TABLE$event)

# Model all proteins
coefficients<-as.data.frame(matrix(ncol= 10))
 
names(coefficients) <- c("AptName","Coeff", "se", "Pvalue", "Cox.zphPVal", "N", "ov.lik.ratio","logrank", "r2", "BIC" )
 
	for (i in 10:ncol(TABLE)) {
	print(colnames(TABLE)[i])
  	protein <- TABLE[,c(i,1:9)]
	j= i-9

#Test Cox model - if there is an error then put NoConverge into results
  	testCox = try(coxph(Surv(protein$time.event, protein$event) ~ protein[,1] + protein$E4_allele + protein$age_baseline + protein$sex + protein$SV1 + protein$SV2 + protein$SV3, data=protein), 
  	              silent = T)
  	
  	if(class(testCox)[1]=="try-error"){
  	  coefficients[j,1]<- paste(colnames(TABLE)[i])
  	  coefficients[j,2:9] <- "NoConverge"
  	} else {
  	
  	#If Cox model is not an error, run and get the coefficients
  	model.cox<- coxph(Surv(protein$time.event, protein$event) ~ protein[,1] + protein$E4_allele + protein$age_baseline + protein$sex + protein$SV1 + protein$SV2 + protein$SV3, data=protein)
  	
  	testkmz <- try(kmz<- cox.zph(model.cox, transform = "km"), silent = T)
  	
  	coefficients[j,1]<- paste(colnames(TABLE)[i])
  	coefficients[j,2]<- summary(model.cox)$coefficients[1,1]
  	coefficients[j,3]<- summary(model.cox)$coefficients[1,3]
  	coefficients[j,4]<- summary(model.cox)$coefficients[1,5]
  	coefficients[j,6]<- model.cox$n
  	coefficients[j,7]<- summary(model.cox)$logtest[[1]]
  	coefficients[j,8]<- summary(model.cox)$sctest[[1]]
  	coefficients[j,9]<- summary(model.cox)$rsq[[1]] # nagelkerke r square
  	coefficients[j,10] <- BIC(model.cox)
  	
  	#If kmz is not an error, get the Cox.zphPVal
  	if(class(testkmz)[1] == "try-error"){
  	coefficients[j,5]<- "solveError"
  	} else {
  	coefficients[j,5]<- kmz$table[1,3]
  	}
  	}
  	}

# Add gene symbols to the coefficients data frame
    prot_list <- getAnalyteInfo(soma_data_replication_PASS) %>% dplyr::select(AptName, EntrezGeneSymbol)
	coefficients$Protein <- prot_list$EntrezGeneSymbol[match(coefficients$AptName,prot_list$AptName)]
	
# FDR adjusted P-values
	coefficients$AdjP <- p.adjust(coefficients$Pvalue, method = "BH")

#Save all output for tables and meta-analysis
fwrite(coefficients,"replication_SomaScan_cph_apoe_all_SVs.txt", quote = F, row.names = F, col.names = T)
```

# Meta-analysis of the discovery and replication cohorts for survival analysis adjusted for APOE status
```{r}
# load the discovery and replication results
discovery_results <- fread("discovery_SomaScan_cph_apoe4_all_SVs.txt") 
replication_results <- fread("replication_SomaScan_cph_apoe_all_SVs.txt")  

discovery_results <- discovery_results %>% dplyr::select(AptName,Coeff,se,Pvalue,N,logrank)
replication_results <- replication_results %>% dplyr::select(AptName,Coeff,se,Pvalue,N,logrank)

colnames(discovery_results) <- c("AptName", "logHR_TPD", "SE_TPD", "Pval_TPD","N_TPD","logrank_TPD")
colnames(replication_results) <- c("AptName", "logHR_OPDC", "SE_OPDC", "Pval_OPDC", "N_OPDC","logrank_OPDC")

# Merge on AptName (common proteins)
meta_data_dementia <- merge(replication_results, discovery_results, by = "AptName", all = TRUE)

meta_data_dementia <- meta_data_dementia %>% drop_na()

meta_data_dementia_long <- reshape(meta_data_dementia, direction = "long",
                          varying = list(c("logHR_OPDC", "logHR_TPD"), 
                                         c("SE_OPDC", "SE_TPD")),
                          v.names = c("logHR", "SE"),
                          times = c("OPDC", "TPD"),
                          timevar = "Study")

meta_results <- list()

for(protein in unique(meta_data_dementia_long$AptName)) {
  
  # Subset data for the current protein
  protein_data <- meta_data_dementia_long[meta_data_dementia_long$AptName == protein, ]
  
  # Run meta-analysis for the current protein
  res <- rma(yi = protein_data$logHR, sei = protein_data$SE, slab = protein_data$Study, data = protein_data)
  
  # Store the result in the list
  meta_results[[protein]] <- res
}

# Create a data frame to store the meta-analysis results for each protein
meta_analysis_summary <- data.frame(
  AptName = character(),
  logHR = numeric(),
  SE = numeric(),
  CI_low = numeric(),
  CI_high = numeric(),
  pval = numeric(),
  I2 = numeric(),
  Tau2 = numeric(),
  Q = numeric(),
  Q_pval = numeric(),  
  stringsAsFactors = FALSE
)

# Loop through the list and extract relevant information for each protein
for(protein in names(meta_results)) {
  
  # Extract the result for the current protein
  res <- meta_results[[protein]]
  
  # Store the summary results in the data frame
  meta_analysis_summary <- rbind(meta_analysis_summary, data.frame(
    AptName = protein,
    logHR = res$b,  # Pooled logHR
    SE = res$se,    # Standard error
    CI_low = res$ci.lb,  # Lower bound of CI
    CI_high = res$ci.ub, # Upper bound of CI
    pval = res$pval,     # p-value
    I2 = res$I2,	           # Heterogeneity I²
    Tau2 = res$tau2,	       # Between-study variance
    Q = res$QE,	            # Q-statistic
    Q_pval = res$QEp       # Q-test p-value	
  ))
}

meta_analysis_summary$Protein <- prot_list$EntrezGeneSymbol[match(meta_analysis_summary$AptName,prot_list$AptName)]

meta_analysis_summary <- meta_analysis_summary %>%
  mutate(Padj = p.adjust(pval, method = "BH"))

write.table(meta_analysis_summary,"meta_pdd_cph_apoe.txt", sep = "\t", quote = F, row.names = F, col.names = T)
```    

# Plot volcano plot for meta-analysis results of survival analysis adjusted for APOE status
```{r}
meta_analysis_summary_apoe <- fread("meta_pdd_cph_apoe.txt") 
pval_threshold <- 0.05

meta_analysis_summary_apoe$color_group <- "grey"  # Default: non-significant
meta_analysis_summary_apoe$color_group[meta_analysis_summary_apoe$Padj < pval_threshold] <- "red"

volcano_plot_survival_meta_apoe <- ggplot(meta_analysis_summary_apoe, aes(x = exp(logHR), y = -log10(Padj), label = Protein, color = color_group)) +
  geom_point(alpha = 0.8, size = 1.3) +
  scale_color_manual(values = c("grey" = "grey", "red" = "red")) +
  geom_hline(yintercept = -log10(pval_threshold), linetype = "dashed", color = "grey") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey") +
  geom_text_repel(data = subset(meta_analysis_summary_apoe, Padj < pval_threshold), size = 3, max.overlaps = 15, color = "black", force = 2, force_pull = 0.5, min.segment.length = 0.2) +
  labs(x = "Hazard Ratio", y = "-log10(FDR)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") + 
  theme(aspect.ratio = 1,
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"),
        legend.position = "none",  # Show legend
        legend.title = element_blank())

ggsave("volcano_plot_survival_apoe_meta.png", 
       volcano_plot_survival_meta_apoe, 
       width = 5, height = 5, dpi = 300)

```

# Plot corres plot for meta-analysis results of survival analysis adjusted for APOE status
```{r}
# Load the discovery and replication results
discovery_results <- fread("discovery_SomaScan_cph_apoe4_all_SVs.txt") 
replication_results <- fread("replication_SomaScan_cph_apoe_all_SVs.txt")  

discovery_results <- discovery_results %>% dplyr::select(AptName,Coeff,se,Pvalue,N,logrank)
replication_results <- replication_results %>% dplyr::select(AptName,Coeff,se,Pvalue,N,logrank)

# Step 1: Find common row names
common_rows <- intersect(discovery_results$AptName, replication_results$AptName)

# Step 2: Subset both data frames to common rows
df1_common <- discovery_results %>% filter(AptName %in% common_rows)
df2_common <- replication_results %>% filter(AptName %in% common_rows)

# Step 3: Merge data
merged <- merge(df1_common, df2_common, by = "AptName", suffixes = c("_disc", "_repl"))
merged$FDR <- meta_analysis_summary_apoe$Padj[match(merged$AptName,meta_analysis_summary_apoe$AptName)]
merged$Gene <- prot_list$EntrezGeneSymbol[match(merged$AptName,prot_list$AptName)]

# Step 4: Identify significant genes
merged <- merged %>% 
   mutate(significant=ifelse(FDR < 0.05, "yes", "no"))

# Step 5: Calculate z-scores of effect sizes for discovery and replication cohorts
merged$z_disc <- merged$Coeff_disc / merged$se_disc
merged$z_repl <- merged$Coeff_repl / merged$se_repl

# Step 6: Create the corres plot
significant_proteins <- subset(merged, significant == "yes")

corres_plot_survival_apoe_z <- ggplot(merged, aes(x = z_disc, y = z_repl)) +
 geom_point(aes(color = significant), size = 1.5, alpha = 0.8) +
 geom_smooth(data = significant_proteins, method = "lm", color = "blue", se = TRUE, fill = "gray") +
  scale_color_manual(values = c("gray", "red")) +
   geom_point(data=subset(merged, significant=="yes"), color="red", size=1.3) + 
 geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.2) +
 geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.2) +
 geom_text_repel(data = subset(merged, significant == "yes"), aes(label = Gene), size = 3, max.overlaps = 15, force_pull = 0.5, min.segment.length = 0.2, force = 2) +
 labs(x = "z-score (Discovery)",
 y = "z-score (Replication)") +
 theme_minimal(base_size = 14) +
 guides(color = "none") +
 theme(aspect.ratio = 1) +
 theme(legend.position = "none") +
 theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
 panel.background = element_blank(), axis.line = element_line(colour = "black"))

ggsave("corres_plot_survival_apoe_z.png", 
       corres_plot_survival_apoe_z , 
       width = 5, height = 5, dpi = 300)
```

# Plot UpSet plot of the meta-analysis results of survival analysis with and without adjustment for APOE status
```{r}
meta_analysis_summary <- fread("meta_pdd_cph.txt")
meta_analysis_summary_apoe <- fread("meta_pdd_cph_apoe.txt")

no_apoe_significant <- meta_analysis_summary %>% filter(Padj < 0.05) #%>% dplyr::select(AptName)
apoe_significant <- meta_analysis_summary_apoe %>% filter(Padj < 0.05) #%>% dplyr::select(AptName)

upset_matrix <- list(
  "Survival_basic" = no_apoe_significant$AptName,
  "Survival_APOE_adjusted" = apoe_significant$AptName
)
  
jpeg("upset_plot_survival_meta.png", res = 150, pointsize = 8)
upset(
  fromList(upset_matrix),
  nsets = 2,  # Number of sets to display 
  nintersects = 30,  # Number of intersections to display (adjust as needed)
  order.by = "freq",  # Order by frequency of intersection
  mainbar.y.label = "Number of Proteins",
  point.size = 3,
  line.size = 1,
  mb.ratio = c(0.6, 0.4)
)
dev.off()
```

# REVIEW
# Pathway enrichement analysis
```{r}
# Load the meta-analysis summary results for survival analysis adjusted for APOE status
meta_analysis_summary_apoe <- fread("meta_pdd_cph_apoe.txt", header = T) 

# Identify significant genes based on adjusted p-value threshold
sig_genes <- meta_analysis_summary_apoe %>%
  filter(Padj < 0.05) %>%
  pull(Protein) %>%
  unique()

# GO annotations
entrez_genes <- bitr(sig_genes, fromType = "SYMBOL",
                     toType = "ENTREZID",
                     OrgDb = org.Hs.eg.db)

ego_all <- enrichGO(gene = entrez_genes$ENTREZID,
                OrgDb = org.Hs.eg.db,
                keyType = "ENTREZID",
                ont = "ALL",     
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.1,
                qvalueCutoff  = 0.2,
                readable = TRUE
                )

# Simplify GO terms
simplified_go_all <- simplify(ego_all,
                           measure = "Wang",  
                           by = "p.adjust",
                           select_fun = min,
                           cutoff = 0.7) 

# KEGG annotations
gene_list <- bitr(sig_genes, 
                  fromType = "SYMBOL", 
                  toType = "ENTREZID",
                  OrgDb = org.Hs.eg.db)

entrez_ids <- gene_list$ENTREZID

kegg_results <- enrichKEGG(gene = entrez_ids,
                           organism = 'hsa',   
                           pAdjustMethod = "BH",
                           pvalueCutoff = 0.05,
                           qvalueCutoff = 0.2)


kegg_results <- kegg_results@result

# REACTOME
reactome <- enrichPathway(
  gene = entrez_genes$ENTREZID,
  pvalueCutoff = 0.1,
  qvalueCutoff = 0.2,
  pAdjustMethod = "BH",
  readable = TRUE
)

# Merge results for GO, KEGG, REACTOME
GO <- simplified_go_all@result %>% dplyr::mutate(Category = paste("GO:",ONTOLOGY, sep=""))
REACT <- reactome@result %>% dplyr::mutate(Category = "REACTOME")

colsToKeep <- colnames(REACT)[which(colnames(REACT) %in% colnames(GO))]
GO <- GO %>% dplyr::select(all_of(colsToKeep))
REACT <- REACT %>% dplyr::select(all_of(colsToKeep))
all <- rbind(GO,REACT)

p_facet_simplified_all <- all %>% 
  filter(p.adjust < 0.05, Count > 1) %>% 
  mutate(log10_padj = -log10(p.adjust)) %>%
  mutate(Description = str_to_sentence(Description)) %>%
  ggplot(aes(x = log10_padj, 
             y = reorder(Description, log10_padj),
             fill = Category)) +
  geom_col() +
  facet_grid(Category ~ ., scales = "free_y", space = "free_y") +
  scale_fill_brewer(palette = "Set2") +
  guides(fill = "none") +
  labs(x = "-log10(P)", y = NULL) +
  theme_bw(base_size = 12) 

ggsave("pathway_enrich_apoe.png", 
       p_facet_simplified_all, 
       width = 8, height = 6, dpi = 300)

```

# StringDB
```{r}
# Load the meta-analysis summary results for survival analysis adjusted for APOE status
meta_analysis_summary_apoe <- fread("meta_pdd_cph_apoe.txt", header = T) 

# Identify significant genes based on adjusted p-value threshold
sig_genes <- meta_analysis_summary_apoe %>%
  filter(Padj < 0.05) %>%
  pull(Protein) %>%
  unique()

# Initialize STRINGdb object with a score threshold of 400
string_db <- STRINGdb$new(version="12.0", species=9606, score_threshold=400)

# Map significant genes to STRING IDs
mapped <- string_db$map(data.frame(Gene=sig_genes), "Gene", removeUnmappedRows=TRUE)

# Map and plot network of protein-protein interactions (PPIs) for the mapped genes
ppi_network <- string_db$get_interactions(mapped$STRING_id)

postscript("string.ps", )
string_db$plot_network(ppi_network)
dev.off()

# Map clusters of protein-protein interactions (PPIs) using the fast greedy algorithm
options(timeout = 300)  
ppi_clusters <- string_db$get_clusters(mapped$STRING_id, algorithm = "fastgreedy")

# Annotate clusters
Pro_clust=data.frame(matrix(NA, nrow = 4, ncol = 21))  
for (i in 1:21) {
  temp=ppi_clusters[[i]]
  temp=unique(mapped$Gene[which(mapped$STRING_id%in%temp)])
  length(temp)=4
  Pro_clust[i]=temp
} 

clusters_list <- apply(Pro_clust, 2, function(x) na.omit(x))
clusters_subset <- clusters_list[1:3]

go_results <- lapply(clusters_subset, function(genes) {
  # Map SYMBOL → ENTREZID
  entrez <- bitr(genes, fromType="SYMBOL", toType="ENTREZID", OrgDb=org.Hs.eg.db)$ENTREZID
  # Skip empty clusters
  if (length(entrez) == 0) return(NULL)
  
  enrichPathway(
  gene = entrez,   # Entrez IDs
  organism = "human",             
  pvalueCutoff = 0.05,            # Set p-value cutoff
  pAdjustMethod = "BH"
  )
})


result=data.frame("ID"=NA,"Description"=NA,"GeneRatio"=NA,"BgRatio"=NA,"pvalue" =NA,"p.adjust"=NA,"qvalue"=NA,"geneID"=NA,"Count"=NA,"Redundant_with_term"=NA,"Category"=NA )

for (i in c(1:3)) {
  genes_full <- bitr(na.omit(Pro_clust[[i]]), fromType ="SYMBOL", toType =  "ENTREZID", OrgDb = org.Hs.eg.db) 
  genes=na.omit(genes_full) 
  genes=genes$ENTREZID  
  ego <- enrichGO(gene = genes,
                  keyType = "ENTREZID", 
                  OrgDb = org.Hs.eg.db,
                  ont = "BP",
                  pAdjustMethod = "BH", 
                  pvalueCutoff  =0.05, 
                  qvalueCutoff  = 0.05,
                  minGSSize     = 1,
                  readable      = TRUE)
  GO_data=data.frame(ego)
  # skip if no enrichment
  if (is.null(ego) || nrow(as.data.frame(ego)) == 0) next
  GO_data=GO_data[order(GO_data$Count,decreasing = T),] 
  
  ego2 <- clusterProfiler::simplify(ego, cutoff=0.7, by="p.adjust", select_fun=min, measure = "Wang", semData = godata('org.Hs.eg.db', ont="BP")) 
  GO_data2=data.frame(ego2)
  # skip again if simplify removed everything
  if (nrow(GO_data2) == 0) next
  GO_data2=GO_data2[order(GO_data2$Count,decreasing = T),] 
  
  # Simplify annotations using GoSemSim
  hsGO <- godata('org.Hs.eg.db', ont="BP") 
  SemSim=mgoSim(GO_data2$ID, GO_data2$ID, semData=hsGO, measure="Wang", combine=NULL)
  GO_data2$Redundant_with_term=NA
  for (j in 1:nrow(GO_data2)) {
    v=SemSim[j,][order(SemSim[j,],decreasing = T)]
    name=rownames(as.data.frame(which(v>=0.25&v!=1))) 
    name=GO_data2$Description[grep(paste0(name,collapse = "|"),GO_data2$ID)]
    GO_data2$Redundant_with_term[j]=paste0(name,collapse = "|")
  }
  GO_data2$GeneRatio=gsub("/","_", GO_data2$GeneRatio)
  GO_data2$BgRatio=gsub("/","_", GO_data2$BgRatio)
  GO_data2$Category=i
  
  if (nrow(result) == 0) {
    result <- GO_data2
  } else {
    missing_cols <- setdiff(colnames(result), colnames(GO_data2))
    for (mc in missing_cols) GO_data2[[mc]] <- NA

    missing_cols2 <- setdiff(colnames(GO_data2), colnames(result))
    for (mc in missing_cols2) result[[mc]] <- NA

    GO_data2 <- GO_data2[, colnames(result)]
    result <- rbind(result, GO_data2)
  }
}
result <- result[-1,]

write.table(result, "string_clusters_go_annotation.txt", sep = "\t", quote = F, row.names = F, col.names = T)
```
