---
title: "Two-sample MR: Proteins to Dementia"
author: "Rafaela Ravazio"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
params:
  clump_kb: null
  clump_r2: null
  outdir_base: null
---

## Overview

This notebook provides a two-sample MR workflow across multiple proteins:

1. Standardize exposure (protein GWAS) and outcome (dementia GWAS) summary statistics.
2. Build per-protein cis instruments with p-value/F-stat filtering and LD clumping.
3. Run MR with harmonization and sensitivity analyses.
4. Optionally run Steiger directionality and MR-PRESSO sensitivity tests.
5. Optionally run cis-window colocalization.
6. Save per-protein outputs and combined summary tables.

## Setup

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(TwoSampleMR)
  library(ggplot2)
  library(coloc)
  library(grid)
})
```

## Configuration

```{r config}
# Returns TRUE when an Rmd YAML parameter is non-null and non-NA (i.e. was
# explicitly passed at render time and should override the cfg default).
has_param_override <- function(value) {
  !is.null(value) && length(value) == 1 && !is.na(value)
}

# Converts a YAML parameter to a finite numeric value, stopping with a clear
# error message if the conversion fails.
get_numeric_param <- function(value, name) {
  value_num <- suppressWarnings(as.numeric(value))
  if (!is.finite(value_num)) {
    stop("Parameter ", name, " must be a single finite numeric value.")
  }
  value_num
}

cfg <- list(
  # Inputs — set these paths to match your local environment.
  exposure_path = "path/to/exposure_gwas.tsv",
  outcome_path  = "path/to/outcome_gwas.tsv",

  # LD reference + PLINK
  ld_bfile  = "path/to/ld_reference/1000G_EUR_chrpos",
  plink_bin = "/path/to/plink",

  # Instrument settings
  clump_kb = 10000,
  clump_r2 = 0.001,
  clump_p = 1,
  p_threshold = 5e-8,
  f_threshold = 10,
  cis_kb = 1000,

  # Harmonization / MR
  harmonise_action = 2, # see TwoSampleMR::harmonise_data
  use_outcome_proxies = FALSE,
  proxy_r2_min = 0.8,
  proxy_window_kb = 1000,

  # Labels
  outcome_name = "Dementia meta-analysis",

  # Steiger directionality (optional)
  run_steiger = TRUE,
  steiger_min_snps = 1,
  steiger_exposure_N = 54219,  # TODO: set to your exposure GWAS N
  steiger_outcome_N = 42140,   # TODO: set to your outcome GWAS N

  # MR-PRESSO (optional)
  run_mr_presso = TRUE,
  mr_presso_min_snps = 4,
  mr_presso_nb_distribution = 1000,
  mr_presso_signif_threshold = 0.05,
  mr_presso_outlier_test = TRUE,
  mr_presso_distortion_test = TRUE,

  # Colocalization
  run_coloc = TRUE,
  outcome_type_coloc = "cc", # "cc" or "quant"
  outcome_N = NA_real_,
  outcome_s = NA_real_,

  # Output
  outdir_base = "./results/two_sample_mr",

  # Reproducibility
  n_workers = 1,
  seed = 42
)

if (exists("params")) {
  if (has_param_override(params$clump_kb)) {
    cfg$clump_kb <- get_numeric_param(params$clump_kb, "clump_kb")
  }
  if (has_param_override(params$clump_r2)) {
    cfg$clump_r2 <- get_numeric_param(params$clump_r2, "clump_r2")
  }
  if (has_param_override(params$outdir_base)) {
    cfg$outdir_base <- as.character(params$outdir_base)
  }
}

# gene_pos values are hg19/GRCh37 coordinates used to define the ±cis_kb window.
protein_settings <- data.table(
  protein = c(
    "HNF4A", "COL28A1", "ASPRV1", "COL6A3", "CD59", "CRABP2", "ARHGAP1", "TBCA", "PPM1A", "PPP2R3A", "RAB14", "DKK3",
    "ERP29", "GPNMB", "RTN4R", "CTSD", "CLSTN3", "PLBD1", "DEFB1", "POSTN", "ST8SIA3", "S100A13", "GLTPD2", "CDH7", "PATE4",
    "CMPK1", "MFAP2", "GAA", "DUT",
    "RIC3", "PVRL3", "LANCL1", "MMP7", "BMP10", "PEX5", "NEGR1", "MAN1B1", "KIAA1024", "IFNGR2", "CTRL", "RBBP6"
  ),
  gene_chr = c(
    20, 7, 2, 2, 11, 1, 11, 5, 14, 3, 9, 11,
    12, 7, 22, 11, 12, 12, 8, 13, 18, 1, 17, 18, 11,
    1, 1, 17, 15,
    11, 3, 2, 11, 2, 12, 1, 9, 15, 21, 16, 16
  ),
  gene_pos = c(
    43063236, 7575504, 70314507, 238322807, 33758025, 156675576, 46722169, 77072163, 60765805, 135866736, 123964147, 12030883,
    112461253, 23314727, 20255841, 1785157, 7311541, 14720817, 6735458, 38172905, 55036162, 153606837, 4693884, 63557573, 125709967,
    47844511, 17308078, 78093680, 48635570,
    8190572, 110913017, 211341483, 102401478, 69098529, 7371170, 72748222, 140003635, 79764646, 34809823, 67965765, 24584184
  )
)

dir.create(cfg$outdir_base, showWarnings = FALSE, recursive = TRUE)
set.seed(cfg$seed)

knitr::kable(protein_settings, caption = "Proteins tested")
```

## Runtime Checks

```{r runtime_checks}
required_files <- c(
  cfg$exposure_path,
  cfg$outcome_path,
  paste0(cfg$ld_bfile, ".bed"),
  paste0(cfg$ld_bfile, ".bim"),
  paste0(cfg$ld_bfile, ".fam")
)

missing_files <- required_files[!file.exists(required_files)]
if (length(missing_files) > 0) {
  stop("Missing required input files:\n", paste(missing_files, collapse = "\n"))
}

if (!file.exists(cfg$plink_bin)) {
  stop("PLINK binary not found at: ", cfg$plink_bin)
}

if (!exists("clump_data", envir = asNamespace("TwoSampleMR"), inherits = FALSE)) {
  stop(
    "TwoSampleMR::clump_data() is not available in this installation.\n",
    "Install or update TwoSampleMR to a version that supports local clumping."
  )
}

clump_formals <- names(formals(get("clump_data", envir = asNamespace("TwoSampleMR"))))
needed_clump_args <- c("clump_kb", "clump_r2", "clump_p1", "bfile", "plink_bin")
missing_clump_args <- setdiff(needed_clump_args, clump_formals)
if (length(missing_clump_args) > 0) {
  stop(
    "Installed TwoSampleMR::clump_data() does not support required local clumping args: ",
    paste(missing_clump_args, collapse = ", "),
    ".\nInstall or update TwoSampleMR to a version with local PLINK support."
  )
}

cfg$n_workers <- suppressWarnings(as.integer(cfg$n_workers[[1]]))
if (!is.finite(cfg$n_workers) || cfg$n_workers < 1L) {
  stop("cfg$n_workers must be a single integer >= 1.")
}

plink_help_ld <- tryCatch(
  system2(cfg$plink_bin, args = c("--help", "ld-snp-list"), stdout = TRUE, stderr = TRUE),
  error = function(e) character()
)
cfg$plink_has_ld_snp_list <- any(grepl("--ld-snp-list", plink_help_ld, fixed = TRUE))

if (cfg$n_workers > 1L && .Platform$OS.type != "unix") {
  warning("cfg$n_workers > 1 requested on a non-Unix platform; falling back to serial execution.")
}

if (isTRUE(cfg$run_mr_presso) && !requireNamespace("MRPRESSO", quietly = TRUE)) {
  stop(
    "MRPRESSO package is required when cfg$run_mr_presso = TRUE.\n",
    "Install with install.packages('MRPRESSO') or set cfg$run_mr_presso = FALSE."
  )
}
```

## Helper Functions

```{r helpers}
# Stops with a clear message if any required columns are absent from a data frame.
must_have <- function(df, cols, nm = "data.frame") {
  miss <- setdiff(cols, names(df))
  if (length(miss)) stop(nm, " missing required columns: ", paste(miss, collapse = ", "))
}

# Reads a PLINK .bim file and normalises chromosome codes (X→23, Y→24, MT→25).
read_bim <- function(bfile) {
  bim_path <- paste0(bfile, ".bim")
  bim <- fread(bim_path, header = FALSE)
  setnames(bim, c("CHR", "SNP", "CM", "POS", "A1", "A2"))

  bim[, CHR := as.character(CHR)]
  bim[CHR %in% c("X", "x"), CHR := "23"]
  bim[CHR %in% c("Y", "y"), CHR := "24"]
  bim[CHR %in% c("MT", "Mt", "mt", "M", "m"), CHR := "25"]
  bim[, CHR := as.integer(CHR)]
  bim
}

# PLINK clumping fails silently when SNP IDs are duplicated in the .bim file.
# This function detects duplicates and writes a deduplicated copy of the bim,
# appending "_dup2", "_dup3", etc. to repeated IDs. The .bed/.fam are reused as-is.
ensure_unique_ld_bfile <- function(bfile, bim, outdir_base, force = FALSE) {
  dup_ids <- bim$SNP[duplicated(bim$SNP)]
  if (length(dup_ids) == 0) {
    message("LD reference has unique SNP IDs.")
    return(list(bfile = bfile, bim = bim, dedup = FALSE))
  }

  message("LD reference has duplicate SNP IDs; creating deduplicated copy.")

  dedup_dir <- file.path(outdir_base, "ld_ref_dedup")
  dir.create(dedup_dir, showWarnings = FALSE, recursive = TRUE)
  out_prefix <- file.path(dedup_dir, paste0(basename(bfile), "_uniqid"))

  if (!force &&
      file.exists(paste0(out_prefix, ".bed")) &&
      file.exists(paste0(out_prefix, ".bim")) &&
      file.exists(paste0(out_prefix, ".fam"))) {
    return(list(bfile = out_prefix, bim = read_bim(out_prefix), dedup = TRUE))
  }

  bim2 <- data.table::as.data.table(bim)
  bim2[, dup_rank := seq_len(.N), by = SNP]
  bim2[dup_rank > 1, SNP := paste0(SNP, "_dup", dup_rank)]

  data.table::fwrite(bim2[, .(CHR, SNP, CM, POS, A1, A2)],
                     paste0(out_prefix, ".bim"),
                     sep = "\t", col.names = FALSE)
  file.copy(paste0(bfile, ".bed"), paste0(out_prefix, ".bed"), overwrite = TRUE)
  file.copy(paste0(bfile, ".fam"), paste0(out_prefix, ".fam"), overwrite = TRUE)

  bim2[, dup_rank := NULL]
  list(bfile = out_prefix, bim = bim2, dedup = TRUE)
}

# Parses variant IDs in CHR:POS or CHR:POS:REF:ALT format (colons or underscores)
# into separate CHR, POS, REF, ALT columns. Used to recover coordinates from
# outcome GWAS files that only provide a variant ID column.
parse_marker <- function(x) {
  x <- as.character(x)
  x2 <- gsub("_", ":", x, fixed = TRUE)
  parts <- tstrsplit(x2, ":", fixed = TRUE)
  out <- data.table(MarkerName = x)

  if (length(parts) >= 2) {
    chr_raw <- gsub("^chr", "", parts[[1]], ignore.case = TRUE)
    chr_raw[chr_raw %in% c("X", "x")] <- "23"
    chr_raw[chr_raw %in% c("Y", "y")] <- "24"
    chr_raw[chr_raw %in% c("MT", "Mt", "mt", "M", "m")] <- "25"
    out[, CHR := suppressWarnings(as.integer(chr_raw))]
    out[, POS := suppressWarnings(as.integer(parts[[2]]))]
  }
  if (length(parts) >= 4) {
    out[, REF := toupper(parts[[3]])]
    out[, ALT := toupper(parts[[4]])]
  }
  out
}

# Returns TRUE for A/T or C/G allele pairs, which are palindromic (strand-ambiguous).
is_palindromic_pair <- function(a1, a2) {
  a1 <- toupper(as.character(a1))
  a2 <- toupper(as.character(a2))

  valid <- !is.na(a1) & !is.na(a2)
  out <- rep(FALSE, length(a1))
  out[valid] <- paste0(a1[valid], a2[valid]) %in% c("AT", "TA", "CG", "GC")
  out
}

# Removes all palindromic SNPs from harmonised data regardless of EAF.
# TwoSampleMR action=2 keeps palindromic SNPs with unambiguous EAF; we drop
# all of them here for conservative analysis (equivalent to action=3).
drop_palindromic_harmonised <- function(dat) {
  keep_attr <- attributes(dat)

  if (!all(c("effect_allele.exposure", "other_allele.exposure") %in% names(dat))) {
    return(dat)
  }

  pal <- is_palindromic_pair(dat$effect_allele.exposure, dat$other_allele.exposure)
  removed <- dat[pal, , drop = FALSE]
  kept <- dat[!pal, , drop = FALSE]

  if (!is.null(keep_attr$audit)) attr(kept, "audit") <- keep_attr$audit
  attr(kept, "palindromic_removed") <- removed
  kept
}

# Returns the first finite numeric value from a vector, or NA if none exist.
first_numeric_or_na <- function(x) {
  x_num <- suppressWarnings(as.numeric(x))
  x_num <- x_num[is.finite(x_num)]
  if (length(x_num) == 0) return(NA_real_)
  x_num[[1]]
}

# Normalises a raw exposure GWAS data frame to a common internal format with
# columns: CHR, POS, ID, effect_allele, other_allele, eaf, beta, se,
# neglog10p, pval, samplesize, PROTEIN, snp, protein_key.
# Handles two input layouts: "legacy" (CHR/POS/ALTFREQ/LOG10P/PROTEIN columns)
# and "new" (VARIANT_ID/chromosome/position/Allele1/Allele2/Effect/StdErr columns).
# Rows with missing coordinates, non-ACGT alleles, or invalid beta/se are dropped.
standardise_exposure <- function(exp_raw) {
  dt <- as.data.table(exp_raw)
  cn <- names(dt)

  is_legacy <- all(c("CHR", "POS", "ID", "REF", "ALT", "ALTFREQ", "N", "BETA", "SE", "LOG10P", "PROTEIN") %in% cn)
  logp_col <- intersect(c("log(P)", "logp", "logP", "LOGP"), cn)
  prot_col <- intersect(c("protein", "PROTEIN", "protein_id"), cn)
  is_new <- all(c("VARIANT_ID", "chromosome", "position", "Allele1", "Allele2", "Effect", "StdErr") %in% cn) &&
    length(logp_col) == 1 && length(prot_col) == 1

  if (!is_legacy && !is_new) {
    stop(
      "Exposure format not recognised. Expected either legacy or new format."
    )
  }

  # Detects whether the log-p column stores -log10(p), ln(p), or log10(p) by
  # correlating each candidate with z-score-derived -log10(p) values, then
  # returns the best-fitting version as -log10(p).
  infer_neglog10p <- function(beta, se, logp_raw) {
    ok <- is.finite(beta) & is.finite(se) & se > 0 & is.finite(logp_raw)

    if (sum(ok) < 1000) {
      med <- median(logp_raw[ok], na.rm = TRUE)
      if (is.finite(med) && med < 0) return(-logp_raw)
      return(logp_raw)
    }

    idx <- which(ok)
    if (length(idx) > 200000) idx <- idx[seq_len(200000)]

    z <- beta[idx] / se[idx]
    pz <- 2 * pnorm(-abs(z))
    target <- -log10(pz)

    c1 <- -logp_raw[idx]
    c2 <- -logp_raw[idx] / log(10)
    c3 <- logp_raw[idx]

    cor1 <- suppressWarnings(cor(c1, target, use = "complete.obs"))
    cor2 <- suppressWarnings(cor(c2, target, use = "complete.obs"))
    cor3 <- suppressWarnings(cor(c3, target, use = "complete.obs"))

    best <- which.max(abs(c(cor1, cor2, cor3)))
    if (best == 1) return(-logp_raw)
    if (best == 2) return(-logp_raw / log(10))
    logp_raw
  }

  if (is_legacy) {
    out <- dt[, .(
      CHR = as.integer(gsub("^chr", "", as.character(CHR), ignore.case = TRUE)),
      POS = as.integer(POS),
      ID = as.character(ID),
      effect_allele = toupper(as.character(ALT)),
      other_allele = toupper(as.character(REF)),
      eaf = suppressWarnings(as.numeric(ALTFREQ)),
      beta = suppressWarnings(as.numeric(BETA)),
      se = suppressWarnings(as.numeric(SE)),
      neglog10p = suppressWarnings(as.numeric(LOG10P)),
      samplesize = suppressWarnings(as.numeric(N)),
      PROTEIN = as.character(PROTEIN)
    )]
  } else {
    lp <- logp_col[[1]]
    pr <- prot_col[[1]]

    out <- dt[, .(
      CHR = as.integer(gsub("^chr", "", as.character(chromosome), ignore.case = TRUE)),
      POS = as.integer(position),
      ID = as.character(VARIANT_ID),
      effect_allele = toupper(as.character(Allele1)),
      other_allele = toupper(as.character(Allele2)),
      eaf = NA_real_,
      beta = suppressWarnings(as.numeric(Effect)),
      se = suppressWarnings(as.numeric(StdErr)),
      logp_raw = suppressWarnings(as.numeric(get(lp))),
      samplesize = NA_real_,
      PROTEIN = as.character(get(pr))
    )]

    out[, neglog10p := infer_neglog10p(beta, se, logp_raw)]
    out[, logp_raw := NULL]
  }

  out[, snp := paste0(CHR, ":", POS)]
  is_acgt <- function(x) !is.na(x) & grepl("^[ACGT]$", x)

  out <- out[
    is.finite(CHR) & is.finite(POS) &
      is.finite(beta) & is.finite(se) & se > 0 &
      is.finite(neglog10p) &
      is_acgt(effect_allele) & is_acgt(other_allele)
  ]

  out[, pval := 10^(-pmin(neglog10p, 300))]
  out[, protein_key := toupper(PROTEIN)]
  out
}

# ---- Instrument diagnostics ----
# These four functions track how many SNPs survive each filter step per protein.
# instrument_diag_init()     creates a one-row data.table of NA counters.
# instrument_diag_set()      updates individual counters in-place.
# instrument_diag_complete() marks the run as ok or failed and returns a copy.
# instrument_diag_write()    combines all proteins, adds boolean survive_* flags,
#                            and writes instrument_filter_counts_all_proteins.tsv.
instrument_diag_init <- function(protein, gene_chr, gene_pos, cfg) {
  data.table(
    protein = protein,
    gene_chr = gene_chr,
    gene_pos = gene_pos,
    cis_start = gene_pos - cfg$cis_kb * 1000,
    cis_end   = gene_pos + cfg$cis_kb * 1000,

    n_exp_protein_total = NA_integer_,
    n_exp_chr_total = NA_integer_,
    n_exp_cis_total = NA_integer_,
    n_exp_after_p = NA_integer_,
    n_exp_with_eaf = NA_integer_,

    n_exp_after_f = NA_integer_,
    n_exp_unique_snp = NA_integer_,
    n_exp_in_ld = NA_integer_,
    n_exp_missing_ld = NA_integer_,
    n_exp_to_clump = NA_integer_,
    n_exp_after_clump = NA_integer_,
    n_outcome_overlap = NA_integer_,
    n_outcome_with_eaf = NA_integer_,

    n_after_format_overlap = NA_integer_,
    n_after_harmonise = NA_integer_,
    n_palindromic_removed = NA_integer_,
    n_mr_keep = NA_integer_,

    status = "started",
    fail_step = NA_character_,
    fail_message = NA_character_
  )
}

instrument_diag_set <- function(diag, ...) {
  vals <- list(...)
  for (nm in names(vals)) {
    data.table::set(diag, j = nm, value = vals[[nm]])
  }
  diag
}

instrument_diag_complete <- function(diag,
                                     status_value = "ok",
                                     fail_step_value = NA_character_,
                                     fail_message_value = NA_character_) {
  data.table::set(diag, j = "status", value = status_value)
  data.table::set(diag, j = "fail_step", value = fail_step_value)
  data.table::set(diag, j = "fail_message", value = fail_message_value)
  data.table::copy(diag)
}

instrument_diag_write <- function(diag_rows,
                                  outdir_base,
                                  file = "instrument_filter_counts_all_proteins.tsv") {
  if (length(diag_rows) == 0L) {
    message("No instrument diagnostics collected.")
    return(invisible(NULL))
  }

  dt <- rbindlist(diag_rows, fill = TRUE, use.names = TRUE)

  dt[, survive_p   := !is.na(n_exp_after_p)   & n_exp_after_p   > 0L]

  dt[, survive_f   := !is.na(n_exp_after_f)   & n_exp_after_f   > 0L]
  dt[, survive_ld := !is.na(n_exp_in_ld) & n_exp_in_ld > 0L]
  dt[, survive_clump := !is.na(n_exp_after_clump) & n_exp_after_clump > 0L]
  dt[, survive_outcome_overlap := !is.na(n_outcome_overlap) & n_outcome_overlap > 0L]
  dt[, survive_harmonise := !is.na(n_after_harmonise) & n_after_harmonise > 0L]
  dt[, survive_mr_keep := !is.na(n_mr_keep) & n_mr_keep > 0L]

  safe_write_tsv(dt, file.path(outdir_base, file))

  invisible(dt)
}

# Keeps one row per SNP ID, retaining the row with the lowest p-value when
# a pval column is present (otherwise the first row encountered).
deduplicate_by_snp <- function(dt) {
  dt <- copy(as.data.table(dt))
  if (!("snp" %in% names(dt))) return(dt)

  if ("pval" %in% names(dt)) {
    setorder(dt, snp, pval)
  } else {
    setorder(dt, snp)
  }
  dt[!duplicated(snp)]
}

# Subsets the global exposure table to SNPs for one protein within the cis window.
subset_exposure_region <- function(exp_dt, protein_key_value, gene_chr, cis_start, cis_end) {
  copy(
    exp_dt[
      protein_key == protein_key_value &
        CHR == gene_chr &
        POS >= cis_start & POS <= cis_end
    ]
  )
}

# Subsets the global outcome table to SNPs within the cis window (used for coloc).
subset_outcome_region <- function(out_dt, gene_chr, cis_start, cis_end) {
  copy(
    out_dt[
      CHR == gene_chr &
        POS >= cis_start & POS <= cis_end
    ]
  )
}

# Returns TRUE only for biallelic SNPs where both alleles are single ACGT bases.
# Filters out indels and multi-character alleles, which cannot be harmonised.
is_acgt_snp <- function(a1, a2) {
  a1 <- toupper(as.character(a1))
  a2 <- toupper(as.character(a2))
  nchar(a1) == 1 & nchar(a2) == 1 & a1 %in% c("A", "C", "G", "T") & a2 %in% c("A", "C", "G", "T")
}

# Normalises a raw outcome GWAS to the same internal format used by the exposure.
# Uses fuzzy column-name matching to handle diverse GWAS file layouts.
# OR and HR columns are log-transformed to beta scale; SE can be derived from CIs.
# Prints the detected column mapping so you can verify the right columns were used.
standardise_outcome <- function(out_raw, verbose = TRUE) {
  dt <- as.data.table(out_raw)
  cn <- names(dt)

  norm <- function(x) gsub("[^a-z0-9]", "", tolower(x))
  cn_norm <- norm(cn)
  pick_col <- function(cands) {
    idx <- match(norm(cands), cn_norm, nomatch = 0)
    if (any(idx > 0)) return(cn[idx[idx > 0][1]])
    NULL
  }

  chr_col <- pick_col(c("chr", "chromosome", "chrom", "chrid"))
  pos_col <- pick_col(c("pos", "position", "basepairlocation", "basepair", "bp", "basepairposition", "base_pair_location"))
  snp_col <- pick_col(c("snp", "variantid", "variant_id", "markername", "marker_name", "rsid", "rs_id"))

  ea_col <- pick_col(c("effectallele", "effect_allele", "allele1", "a1", "alt", "alt_allele"))
  oa_col <- pick_col(c("otherallele", "other_allele", "noneffect_allele", "noneffectallele", "allele2", "a2", "ref", "ref_allele"))

  beta_col <- pick_col(c("beta", "effect", "estimate", "logodds", "logor", "log_or", "loghr", "log_hr"))
  se_col <- pick_col(c("se", "standarderror", "stderr", "stderror", "sebeta", "se_beta", "stderrbeta"))
  pval_col <- pick_col(c("pval", "pvalue", "p_value", "p-value", "p"))
  eaf_col <- pick_col(c("eaf", "effectallelefrequency", "effect_allele_frequency", "freq1", "allele1freq", "allele1frequency"))
  n_col <- pick_col(c(
    "n", "samplesize", "sample_size", "samplesizetotal", "totalsamplesize", "total_sample_size",
    "totaln", "n_total", "n_total_samples", "neff"
  ))
  or_col <- pick_col(c("oddsratio", "odds_ratio", "or"))
  hr_col <- pick_col(c("hr", "hazardratio", "hazard_ratio"))
  ci_l_col <- pick_col(c("cilower", "ci_lower", "lowerci", "lower_ci", "lci", "ci_lower95", "lowerci95"))
  ci_u_col <- pick_col(c("ciupper", "ci_upper", "upperci", "upper_ci", "uci", "ci_upper95", "upperci95"))

  if (is.null(ea_col) || is.null(oa_col)) {
    stop("Outcome format not recognised: missing effect/other allele columns.")
  }
  if (is.null(beta_col) && is.null(or_col) && is.null(hr_col)) {
    stop("Outcome format not recognised: need beta or OR/HR columns.")
  }
  if (is.null(se_col) && (is.null(ci_l_col) || is.null(ci_u_col))) {
    stop("Outcome format not recognised: need SE or confidence intervals.")
  }

  out <- data.table()
  out[, effect_allele := toupper(as.character(dt[[ea_col]]))]
  out[, other_allele := toupper(as.character(dt[[oa_col]]))]

  if (!is.null(chr_col) && !is.null(pos_col)) {
    out[, CHR := as.integer(gsub("^chr", "", as.character(dt[[chr_col]]), ignore.case = TRUE))]
    out[, POS := as.integer(dt[[pos_col]])]
  } else if (!is.null(snp_col)) {
    parsed <- parse_marker(dt[[snp_col]])
    out[, CHR := parsed$CHR]
    out[, POS := parsed$POS]
  } else {
    out[, `:=`(CHR = NA_integer_, POS = NA_integer_)]
  }

  snp_raw <- if (!is.null(snp_col)) as.character(dt[[snp_col]]) else NA_character_
  out[, snp := fifelse(is.finite(CHR) & is.finite(POS), paste0(CHR, ":", POS), snp_raw)]

  out[, beta := NA_real_]
  if (!is.null(beta_col)) out[, beta := suppressWarnings(as.numeric(dt[[beta_col]]))]

  if (!is.null(or_col)) {
    orv <- suppressWarnings(as.numeric(dt[[or_col]]))
    out[!is.finite(beta) & is.finite(orv) & orv > 0, beta := log(orv)]
  }
  if (!is.null(hr_col)) {
    hrv <- suppressWarnings(as.numeric(dt[[hr_col]]))
    out[!is.finite(beta) & is.finite(hrv) & hrv > 0, beta := log(hrv)]
  }

  out[, se := NA_real_]
  if (!is.null(se_col)) out[, se := suppressWarnings(as.numeric(dt[[se_col]]))]

  if (all(!is.finite(out$se)) && !is.null(ci_l_col) && !is.null(ci_u_col)) {
    lo <- suppressWarnings(as.numeric(dt[[ci_l_col]]))
    hi <- suppressWarnings(as.numeric(dt[[ci_u_col]]))
    if (!is.null(or_col) || !is.null(hr_col)) {
      lo <- ifelse(is.finite(lo) & lo > 0, log(lo), NA_real_)
      hi <- ifelse(is.finite(hi) & hi > 0, log(hi), NA_real_)
    }
    out[, se := (hi - lo) / (2 * 1.96)]
  }

  out[, pval := if (!is.null(pval_col)) suppressWarnings(as.numeric(dt[[pval_col]])) else NA_real_]
  out[, eaf := if (!is.null(eaf_col)) suppressWarnings(as.numeric(dt[[eaf_col]])) else NA_real_]
  out[, samplesize := if (!is.null(n_col)) suppressWarnings(as.numeric(dt[[n_col]])) else NA_real_]
  out[, TotalSampleSize := samplesize]

  if (all(!is.finite(out$pval)) && any(is.finite(out$beta)) && any(is.finite(out$se))) {
    z <- out$beta / out$se
    out[, pval := 2 * pnorm(-abs(z))]
  }

  if (verbose) {
    message("Outcome columns used: ",
            "CHR=", ifelse(is.null(chr_col), "NA", chr_col), ", ",
            "POS=", ifelse(is.null(pos_col), "NA", pos_col), ", ",
            "SNP=", ifelse(is.null(snp_col), "NA", snp_col), ", ",
            "EA=", ea_col, ", OA=", oa_col, ", ",
            "BETA=", ifelse(is.null(beta_col), "NA", beta_col), ", ",
            "SE=", ifelse(is.null(se_col), "NA", se_col), ", ",
            "P=", ifelse(is.null(pval_col), "NA", pval_col), ", ",
            "EAF=", ifelse(is.null(eaf_col), "NA", eaf_col), ", ",
            "N=", ifelse(is.null(n_col), "NA", n_col))
  }

  out <- out[!is.na(snp) & is.finite(beta) & is.finite(se) & se > 0]
  out
}

# Wraps TwoSampleMR::format_data to convert our internal format into the
# column-naming convention expected by harmonise_data() and mr().
# Only passes optional columns (eaf, samplesize, chr, pos) when they are present.
format_tsmr_data <- function(df, type, label) {
  args <- list(
    dat = as.data.frame(df),
    type = type,
    snp_col = "snp",
    beta_col = "beta",
    se_col = "se",
    effect_allele_col = "effect_allele",
    other_allele_col = "other_allele"
  )

  if ("pval" %in% names(df) && any(is.finite(df$pval))) args$pval_col <- "pval"
  if ("eaf" %in% names(df) && any(is.finite(df$eaf))) args$eaf_col <- "eaf"
  if ("samplesize" %in% names(df) && any(is.finite(df$samplesize))) args$samplesize_col <- "samplesize"
  if (all(c("CHR", "POS") %in% names(df)) && any(is.finite(df$CHR)) && any(is.finite(df$POS))) {
    args$chr_col <- "CHR"
    args$pos_col <- "POS"
  }

  out <- do.call(TwoSampleMR::format_data, args)
  if (type == "exposure") {
    out$exposure <- label
    out$id.exposure <- label
  } else {
    out$outcome <- label
    out$id.outcome <- label
  }
  out
}

# Runs LD clumping via local PLINK using TwoSampleMR::clump_data().
# Saves the pre- and post-clump data to outdir for auditing.
# Returns the subset of exp_df rows corresponding to the lead (index) SNPs.
clump_exposure_local_tsmr <- function(exp_df, cfg, ld_bfile, protein_of_interest, outdir) {
  exp_fmt <- as.data.table(format_tsmr_data(exp_df, type = "exposure", label = protein_of_interest))

  audit_input <- file.path(outdir, paste0(protein_of_interest, "_exp_for_clump_tsmr.tsv"))
  safe_write_tsv(exp_fmt, audit_input)

  clump_args <- list(
    dat = as.data.frame(exp_fmt),
    clump_kb = cfg$clump_kb,
    clump_r2 = cfg$clump_r2,
    clump_p1 = cfg$clump_p,
    bfile = ld_bfile,
    plink_bin = cfg$plink_bin
  )

  exp_clumped <- tryCatch(
    as.data.table(do.call(TwoSampleMR::clump_data, clump_args)),
    error = function(e) {
      stop("TwoSampleMR clumping failed: ", e$message, call. = FALSE)
    }
  )

  audit_output <- file.path(outdir, paste0(protein_of_interest, "_exp_clumped_tsmr.tsv"))
  safe_write_tsv(exp_clumped, audit_output)

  must_have(exp_clumped, "SNP", "TwoSampleMR clumped output")

  lead_snps <- unique(trimws(as.character(exp_clumped$SNP)))
  lead_snps <- lead_snps[!is.na(lead_snps) & lead_snps != ""]
  if (length(lead_snps) == 0) {
    stop("TwoSampleMR clumping produced no lead SNPs.")
  }

  exp_keep <- exp_df[snp %in% lead_snps]
  if (nrow(exp_keep) == 0) {
    stop("TwoSampleMR clumping returned lead SNPs that could not be mapped back to exposure instruments.")
  }

  exp_keep
}

# Writes a data.table to a tab-separated file, creating the directory if needed.
safe_write_tsv <- function(dt, path) {
  dir.create(dirname(path), recursive = TRUE, showWarnings = FALSE)
  fwrite(as.data.table(dt), path, sep = "\t")
}

# Returns the complementary DNA base for each element (A↔T, C↔G).
complement_alleles <- function(x) {
  comp_base <- c(A = "T", T = "A", C = "G", G = "C")
  x <- toupper(as.character(x))
  out <- unname(comp_base[x])
  out[is.na(out)] <- NA_character_
  out
}

# Checks whether a SNP's alleles from the GWAS match those in the LD reference,
# allowing for strand flips (complement) and allele ordering swaps.
# Returns list(ok=TRUE, ld_effect, ld_other, mode) on success, or
# list(ok=FALSE, reason) if alleles cannot be reconciled.
align_dataset_alleles_to_ld <- function(effect_allele, other_allele, ld_a1, ld_a2) {
  ea <- toupper(as.character(effect_allele))
  oa <- toupper(as.character(other_allele))
  ld_a1 <- toupper(as.character(ld_a1))
  ld_a2 <- toupper(as.character(ld_a2))

  valid <- c(ea, oa, ld_a1, ld_a2)
  if (any(is.na(valid)) || any(!grepl("^[ACGT]$", valid))) {
    return(list(ok = FALSE, reason = "non_acgt_or_missing"))
  }

  if (ea == ld_a1 && oa == ld_a2) {
    return(list(ok = TRUE, ld_effect = ld_a1, ld_other = ld_a2, mode = "as_is"))
  }
  if (ea == ld_a2 && oa == ld_a1) {
    return(list(ok = TRUE, ld_effect = ld_a2, ld_other = ld_a1, mode = "swap"))
  }

  c_a1 <- complement_alleles(ld_a1)
  c_a2 <- complement_alleles(ld_a2)
  if (ea == c_a1 && oa == c_a2) {
    return(list(ok = TRUE, ld_effect = ld_a1, ld_other = ld_a2, mode = "complement"))
  }
  if (ea == c_a2 && oa == c_a1) {
    return(list(ok = TRUE, ld_effect = ld_a2, ld_other = ld_a1, mode = "complement+swap"))
  }

  list(ok = FALSE, reason = "alleles_do_not_match_ld")
}

# Parses a PLINK phase string (e.g. "AG/TC") into the two haplotypes.
# target_is_a=TRUE means the target SNP is SNP_A in the PLINK output; the
# first character of each haplotype is the target allele, the second is proxy.
parse_plink_phase <- function(phase, target_is_a) {
  parts <- strsplit(as.character(phase), "/", fixed = TRUE)[[1]]
  if (length(parts) != 2L || any(nchar(parts) != 2L)) return(NULL)

  hap1 <- strsplit(parts[[1]], "", fixed = TRUE)[[1]]
  hap2 <- strsplit(parts[[2]], "", fixed = TRUE)[[1]]

  if (isTRUE(target_is_a)) {
    list(
      target = c(hap1[[1]], hap2[[1]]),
      proxy = c(hap1[[2]], hap2[[2]])
    )
  } else {
    list(
      target = c(hap1[[2]], hap2[[2]]),
      proxy = c(hap1[[1]], hap2[[1]])
    )
  }
}

# Uses the PLINK phase string to determine whether the proxy SNP's effect
# direction should be flipped relative to the target SNP.
# A flip is needed when the proxy effect allele is on the opposite haplotype
# from the target effect allele.
infer_proxy_flip_from_phase <- function(target_ld_effect, proxy_ld_effect, phase, target_is_a) {
  parsed <- parse_plink_phase(phase, target_is_a = target_is_a)
  if (is.null(parsed)) return(list(ok = FALSE, reason = "invalid_phase"))

  target_idx <- which(parsed$target == target_ld_effect)
  if (length(target_idx) != 1L) {
    return(list(ok = FALSE, reason = "ambiguous_target_phase"))
  }

  if (parsed$proxy[target_idx] == proxy_ld_effect) {
    return(list(ok = TRUE, flip = FALSE))
  }

  other_idx <- if (target_idx == 1L) 2L else 1L
  if (parsed$proxy[other_idx] == proxy_ld_effect) {
    return(list(ok = TRUE, flip = TRUE))
  }

  list(ok = FALSE, reason = "proxy_effect_not_present_in_phase")
}

# Returns a zero-row data.table with the expected schema for proxy search results,
# used as a safe initialiser before rbindlist accumulation.
empty_proxy_search_results <- function() {
  data.table(
    target_snp = character(),
    proxy_snp = character(),
    r2 = numeric(),
    phase = character(),
    target_is_a = logical(),
    in_outcome = logical(),
    ld_chr_a = integer(),
    ld_bp_a = integer(),
    ld_chr_b = integer(),
    ld_bp_b = integer()
  )
}

# Reads the .ld file produced by PLINK --r2 and extracts target/proxy SNP pairs
# with their r², phase string, and genomic positions.
parse_plink_proxy_output <- function(ld_path, target_snps) {
  if (!file.exists(ld_path) || file.info(ld_path)$size == 0) {
    return(empty_proxy_search_results())
  }

  ld <- suppressWarnings(fread(ld_path))
  if (nrow(ld) == 0) return(empty_proxy_search_results())
  setnames(ld, trimws(names(ld)))

  must_have(ld, c("SNP_A", "SNP_B", "CHR_A", "CHR_B", "BP_A", "BP_B", "PHASE"), "PLINK LD output")
  r2_col <- intersect(c("R2", "R^2", "R2.1"), names(ld))
  if (length(r2_col) == 0) stop("PLINK LD output missing R2 column.")

  ld <- as.data.table(ld)
  target_snps <- unique(as.character(target_snps))
  ld[, `:=`(
    SNP_A = as.character(SNP_A),
    SNP_B = as.character(SNP_B),
    r2 = suppressWarnings(as.numeric(get(r2_col[[1]])))
  )]
  ld <- ld[is.finite(r2)]
  if (nrow(ld) == 0) return(empty_proxy_search_results())

  ld <- ld[
    (SNP_A %chin% target_snps & SNP_B != SNP_A) |
      (SNP_B %chin% target_snps & SNP_A != SNP_B)
  ]
  if (nrow(ld) == 0) return(empty_proxy_search_results())

  ld[, target_snp := fifelse(SNP_A %chin% target_snps, SNP_A, SNP_B)]
  ld[, target_is_a := SNP_A == target_snp]
  ld[, proxy_snp := fifelse(target_is_a, SNP_B, SNP_A)]
  ld[, `:=`(
    phase = as.character(PHASE),
    ld_chr_a = suppressWarnings(as.integer(CHR_A)),
    ld_bp_a = suppressWarnings(as.integer(BP_A)),
    ld_chr_b = suppressWarnings(as.integer(CHR_B)),
    ld_bp_b = suppressWarnings(as.integer(BP_B))
  )]

  unique(ld[, .(
    target_snp,
    proxy_snp,
    r2,
    phase,
    target_is_a,
    in_outcome = FALSE,
    ld_chr_a,
    ld_bp_a,
    ld_chr_b,
    ld_bp_b
  )])
}

# Calls PLINK --r2 for a single target SNP to find all LD proxies within
# cfg$proxy_window_kb, returning results via parse_plink_proxy_output().
run_plink_proxy_search <- function(target_snp, cfg, ld_bfile) {
  out_prefix <- tempfile(
    pattern = paste0("proxy_ld_", gsub("[^A-Za-z0-9]+", "_", target_snp), "_"),
    tmpdir = tempdir()
  )

  args <- c(
    "--bfile", ld_bfile,
    "--r2", "in-phase", "with-freqs",
    "--ld-snp", target_snp,
    "--ld-window", "999999",
    "--ld-window-kb", as.character(cfg$proxy_window_kb),
    "--ld-window-r2", "0",
    "--out", out_prefix
  )

  plink_log <- tryCatch(
    system2(cfg$plink_bin, args = args, stdout = TRUE, stderr = TRUE),
    error = function(e) {
      stop("PLINK proxy search failed for ", target_snp, ": ", e$message, call. = FALSE)
    }
  )
  status <- attr(plink_log, "status")
  if (!is.null(status) && status != 0) {
    stop(
      "PLINK proxy search failed for ", target_snp, ": ",
      paste(tail(plink_log, 20), collapse = "\n"),
      call. = FALSE
    )
  }

  parse_plink_proxy_output(paste0(out_prefix, ".ld"), target_snp)
}

# Batched version of run_plink_proxy_search() for multiple target SNPs.
# Uses PLINK --ld-snp-list when available (one call for all SNPs); otherwise
# falls back to looping run_plink_proxy_search() one SNP at a time.
run_plink_proxy_search_many <- function(target_snps, cfg, ld_bfile) {
  target_snps <- unique(as.character(target_snps))
  target_snps <- target_snps[!is.na(target_snps) & nzchar(target_snps)]
  if (length(target_snps) == 0L) {
    return(empty_proxy_search_results())
  }
  if (length(target_snps) == 1L || !isTRUE(cfg$plink_has_ld_snp_list)) {
    return(rbindlist(lapply(target_snps, run_plink_proxy_search, cfg = cfg, ld_bfile = ld_bfile), fill = TRUE))
  }

  out_prefix <- tempfile(pattern = "proxy_ld_batch_", tmpdir = tempdir())
  snp_list_path <- paste0(out_prefix, "_targets.txt")
  writeLines(target_snps, con = snp_list_path)

  args <- c(
    "--bfile", ld_bfile,
    "--r2", "in-phase", "with-freqs",
    "--ld-snp-list", snp_list_path,
    "--ld-window", "999999",
    "--ld-window-kb", as.character(cfg$proxy_window_kb),
    "--ld-window-r2", "0",
    "--out", out_prefix
  )

  plink_log <- tryCatch(
    system2(cfg$plink_bin, args = args, stdout = TRUE, stderr = TRUE),
    error = function(e) {
      stop("PLINK batched proxy search failed: ", e$message, call. = FALSE)
    }
  )
  status <- attr(plink_log, "status")
  if (!is.null(status) && status != 0) {
    stop(
      "PLINK batched proxy search failed: ",
      paste(tail(plink_log, 20), collapse = "\n"),
      call. = FALSE
    )
  }

  parse_plink_proxy_output(paste0(out_prefix, ".ld"), target_snps)
}

# Constructs a synthetic outcome row for a target SNP using a proxy SNP's
# summary statistics. Aligns alleles via the LD reference, then applies a
# sign flip to beta (and 1-eaf to EAF) if the phase indicates the proxy
# effect allele is on the opposite haplotype from the target effect allele.
build_proxy_outcome_row <- function(target_row, proxy_outcome_row, ld_hit, bim) {
  target_bim <- bim[SNP == target_row$snp]
  proxy_bim <- bim[SNP == proxy_outcome_row$snp]

  if (nrow(target_bim) != 1L) {
    return(list(ok = FALSE, reason = "target_ld_record_missing_or_duplicated"))
  }
  if (nrow(proxy_bim) != 1L) {
    return(list(ok = FALSE, reason = "proxy_ld_record_missing_or_duplicated"))
  }

  target_map <- align_dataset_alleles_to_ld(
    effect_allele = target_row$effect_allele,
    other_allele = target_row$other_allele,
    ld_a1 = target_bim$A1[[1]],
    ld_a2 = target_bim$A2[[1]]
  )
  if (!isTRUE(target_map$ok)) return(list(ok = FALSE, reason = paste0("target_", target_map$reason)))

  proxy_map <- align_dataset_alleles_to_ld(
    effect_allele = proxy_outcome_row$effect_allele,
    other_allele = proxy_outcome_row$other_allele,
    ld_a1 = proxy_bim$A1[[1]],
    ld_a2 = proxy_bim$A2[[1]]
  )
  if (!isTRUE(proxy_map$ok)) return(list(ok = FALSE, reason = paste0("proxy_", proxy_map$reason)))

  flip_info <- infer_proxy_flip_from_phase(
    target_ld_effect = target_map$ld_effect,
    proxy_ld_effect = proxy_map$ld_effect,
    phase = ld_hit$phase[[1]],
    target_is_a = ld_hit$target_is_a[[1]]
  )
  if (!isTRUE(flip_info$ok)) return(list(ok = FALSE, reason = flip_info$reason))

  proxy_snp <- as.character(proxy_outcome_row$snp[[1]])
  syn <- copy(proxy_outcome_row)
  syn[, proxy_target_snp := target_row$snp[[1]]]
  syn[, proxy_snp := proxy_snp]
  syn[, proxy_r2 := ld_hit$r2[[1]]]
  syn[, proxy_phase := ld_hit$phase[[1]]]
  syn[, proxy_beta_flipped := flip_info$flip]
  syn[, proxy_source := "local_plink_ld"]

  syn[, snp := target_row$snp[[1]]]
  syn[, CHR := target_row$CHR[[1]]]
  syn[, POS := target_row$POS[[1]]]
  syn[, effect_allele := target_row$effect_allele[[1]]]
  syn[, other_allele := target_row$other_allele[[1]]]
  syn[, beta := if (isTRUE(flip_info$flip)) -beta else beta]
  if ("eaf" %in% names(syn)) {
    syn[, eaf := ifelse(is.finite(eaf) & isTRUE(flip_info$flip), 1 - eaf, eaf)]
  }

  list(ok = TRUE, row = syn, flip = flip_info$flip)
}

# For each exposure instrument missing from the outcome GWAS, searches for an
# LD proxy (r² ≥ cfg$proxy_r2_min within cfg$proxy_window_kb) and substitutes
# it using build_proxy_outcome_row(). Returns status per SNP and a combined
# outcome table with proxies inserted in place of unmatched targets.
rescue_outcome_proxies <- function(target_snps, exp_df, out_best, bim, cfg, ld_bfile) {
  target_snps <- unique(as.character(target_snps))
  exp_lookup <- copy(as.data.table(exp_df))
  setindex(exp_lookup, snp)

  out_best <- copy(as.data.table(out_best))
  setindex(out_best, snp)

  native_snps <- unique(out_best$snp[out_best$snp %chin% target_snps])
  available_outcome_snps <- unique(out_best$snp)

  status_dt <- data.table(target_snp = target_snps)
  status_dt[, `:=`(
    status = fifelse(target_snp %in% native_snps, "found_natively", NA_character_),
    proxy_snp = NA_character_,
    proxy_r2 = NA_real_,
    proxy_beta_flipped = as.logical(NA),
    message = NA_character_
  )]

  search_results <- empty_proxy_search_results()
  rescued_rows <- list()

  missing <- setdiff(target_snps, native_snps)
  ld_hits_all <- empty_proxy_search_results()

  if (length(missing) > 0L) {
    ld_hits_all <- tryCatch(
      run_plink_proxy_search_many(missing, cfg = cfg, ld_bfile = ld_bfile),
      error = function(e) {
        for (target_key in missing) {
          ld_hits <- tryCatch(
            run_plink_proxy_search(target_key, cfg = cfg, ld_bfile = ld_bfile),
            error = function(e_single) {
              status_dt[target_snp == target_key, `:=`(
                status = "proxy_search_failed",
                message = e_single$message
              )]
              NULL
            }
          )
          if (!is.null(ld_hits)) {
            ld_hits_all <<- rbindlist(list(ld_hits_all, ld_hits), fill = TRUE, use.names = TRUE)
          }
        }
        ld_hits_all
      }
    )

    if (nrow(ld_hits_all) > 0L) {
      ld_hits_all[, in_outcome := proxy_snp %chin% available_outcome_snps]
      search_results <- rbindlist(list(search_results, ld_hits_all), fill = TRUE, use.names = TRUE)
    }
  }

  for (target_key in missing) {
    ld_hits <- ld_hits_all[target_snp == target_key]
    if (nrow(ld_hits) == 0) {
      if (is.na(status_dt[target_snp == target_key, status][[1]])) {
        status_dt[target_snp == target_key, `:=`(
          status = "not_found_in_outcome",
          message = "No LD proxy candidates found in search window."
        )]
      }
      next
    }

    outcome_hits <- ld_hits[in_outcome == TRUE]
    if (nrow(outcome_hits) == 0) {
      status_dt[target_snp == target_key, `:=`(
        status = "not_found_in_outcome",
        message = "No LD proxy candidates were present in the outcome data."
      )]
      next
    }

    setorder(outcome_hits, -r2, proxy_snp)
    if (!any(outcome_hits$r2 >= cfg$proxy_r2_min, na.rm = TRUE)) {
      best_below <- outcome_hits[1]
      status_dt[target_snp == target_key, `:=`(
        status = "proxy_below_threshold",
        proxy_snp = best_below$proxy_snp,
        proxy_r2 = best_below$r2,
        message = paste0("Best outcome proxy below r2 threshold of ", cfg$proxy_r2_min, ".")
      )]
      next
    }

    best_hit <- outcome_hits[r2 >= cfg$proxy_r2_min][1]
    target_row <- exp_lookup[snp == target_key][1]
    proxy_row <- out_best[snp == best_hit$proxy_snp][1]
    built <- build_proxy_outcome_row(
      target_row = target_row,
      proxy_outcome_row = proxy_row,
      ld_hit = best_hit,
      bim = bim
    )

    if (!isTRUE(built$ok)) {
      status_dt[target_snp == target_key, `:=`(
        status = "proxy_ambiguous_phase_or_alleles",
        proxy_snp = best_hit$proxy_snp,
        proxy_r2 = best_hit$r2,
        message = built$reason
      )]
      next
    }

    rescued_rows[[length(rescued_rows) + 1L]] <- built$row
    status_dt[target_snp == target_key, `:=`(
      status = "rescued_by_proxy",
      proxy_snp = best_hit$proxy_snp,
      proxy_r2 = best_hit$r2,
      proxy_beta_flipped = isTRUE(built$flip),
      message = "Proxy substituted from local PLINK LD search."
    )]
  }

  rescued_dt <- if (length(rescued_rows) > 0L) {
    rbindlist(rescued_rows, fill = TRUE, use.names = TRUE)
  } else {
    out_best[0]
  }
  unresolved_dt <- status_dt[status %in% c("not_found_in_outcome", "proxy_below_threshold", "proxy_ambiguous_phase_or_alleles", "proxy_search_failed")]

  list(
    status = status_dt,
    search_results = search_results,
    rescued_rows = rescued_dt,
    unresolved = unresolved_dt
  )
}

# Runs the Steiger directionality test to check that the protein (exposure)
# explains more variance in its own GWAS than in the outcome GWAS, confirming
# the assumed causal direction. Requires sample sizes; falls back to
# cfg$steiger_exposure_N / cfg$steiger_outcome_N when not in the harmonised data.
run_steiger_directionality <- function(d1, cfg, protein_of_interest) {
  dt <- as.data.table(d1)
  if (nrow(dt) < cfg$steiger_min_snps) return(NULL)

  dt[, pval.exposure := suppressWarnings(as.numeric(pval.exposure))]
  dt[, pval.outcome := suppressWarnings(as.numeric(pval.outcome))]

  bad_px <- !is.finite(dt$pval.exposure)
  bad_py <- !is.finite(dt$pval.outcome)
  if (any(bad_px)) dt[bad_px, pval.exposure := 2 * pnorm(-abs(beta.exposure / se.exposure))]
  if (any(bad_py)) dt[bad_py, pval.outcome := 2 * pnorm(-abs(beta.outcome / se.outcome))]

  if (!("samplesize.exposure" %in% names(dt))) dt[, samplesize.exposure := NA_real_]
  if (!("samplesize.outcome" %in% names(dt))) dt[, samplesize.outcome := NA_real_]
  dt[, samplesize.exposure := suppressWarnings(as.numeric(samplesize.exposure))]
  dt[, samplesize.outcome := suppressWarnings(as.numeric(samplesize.outcome))]

  if (is.finite(cfg$steiger_exposure_N)) dt[!is.finite(samplesize.exposure), samplesize.exposure := cfg$steiger_exposure_N]
  if (is.finite(cfg$steiger_outcome_N)) dt[!is.finite(samplesize.outcome), samplesize.outcome := cfg$steiger_outcome_N]

  if (any(is.finite(dt$samplesize.exposure))) {
    n_exp <- median(dt$samplesize.exposure[is.finite(dt$samplesize.exposure)], na.rm = TRUE)
    dt[!is.finite(samplesize.exposure), samplesize.exposure := n_exp]
  }
  if (any(is.finite(dt$samplesize.outcome))) {
    n_out <- median(dt$samplesize.outcome[is.finite(dt$samplesize.outcome)], na.rm = TRUE)
    dt[!is.finite(samplesize.outcome), samplesize.outcome := n_out]
  }

  dt <- dt[
    is.finite(pval.exposure) &
      is.finite(pval.outcome) &
      is.finite(samplesize.exposure) &
      is.finite(samplesize.outcome)
  ]
  if (nrow(dt) < cfg$steiger_min_snps) return(NULL)

  res <- tryCatch(
    TwoSampleMR::directionality_test(as.data.frame(dt)),
    error = function(e) {
      message("Steiger failed for ", protein_of_interest, ": ", e$message)
      NULL
    }
  )
  if (is.null(res) || nrow(res) == 0) return(NULL)

  out <- as.data.table(res)
  out[, `:=`(
    protein = protein_of_interest,
    nsnp_steiger = nrow(dt)
  )]
  out
}

# Runs MR-PRESSO to detect and correct for horizontal pleiotropy outliers.
# Requires ≥ cfg$mr_presso_min_snps instruments. Returns a list with the
# main MR estimates, global test p-value, per-SNP outlier test, and
# distortion test (comparing corrected vs uncorrected estimates).
run_mr_presso_test <- function(d1, cfg, protein_of_interest) {
  dt <- as.data.table(d1)
  dt <- dt[
    is.finite(beta.outcome) & is.finite(beta.exposure) &
      is.finite(se.outcome) & is.finite(se.exposure) &
      se.outcome > 0 & se.exposure > 0
  ]

  if (nrow(dt) < cfg$mr_presso_min_snps) return(NULL)

  presso_res <- tryCatch(
    MRPRESSO::mr_presso(
      BetaOutcome = "beta.outcome",
      BetaExposure = "beta.exposure",
      SdOutcome = "se.outcome",
      SdExposure = "se.exposure",
      OUTLIERtest = isTRUE(cfg$mr_presso_outlier_test),
      DISTORTIONtest = isTRUE(cfg$mr_presso_distortion_test),
      data = as.data.frame(dt),
      NbDistribution = as.integer(cfg$mr_presso_nb_distribution),
      SignifThreshold = as.numeric(cfg$mr_presso_signif_threshold)
    ),
    error = function(e) {
      message("MR-PRESSO failed for ", protein_of_interest, ": ", e$message)
      NULL
    }
  )
  if (is.null(presso_res)) return(NULL)

  get_comp <- function(obj, nm) {
    if (!is.null(obj[[nm]])) return(obj[[nm]])
    if (!is.null(obj[["MR-PRESSO results"]]) && !is.null(obj[["MR-PRESSO results"]][[nm]])) {
      return(obj[["MR-PRESSO results"]][[nm]])
    }
    NULL
  }

  as_dt <- function(x, row_col = NULL) {
    if (is.null(x)) return(NULL)
    if (is.data.frame(x)) {
      if (!is.null(row_col)) return(as.data.table(x, keep.rownames = row_col))
      return(as.data.table(x))
    }
    if (is.atomic(x) && length(x) > 0) return(as.data.table(as.list(x)))
    if (is.list(x)) return(as.data.table(x))
    NULL
  }

  main_dt <- as_dt(get_comp(presso_res, "Main MR results"), row_col = "method")
  global_dt <- as_dt(get_comp(presso_res, "Global Test"))
  outlier_dt <- as_dt(get_comp(presso_res, "Outlier Test"), row_col = "SNP")
  distortion_raw <- get_comp(presso_res, "Distortion Test")

  distortion_dt <- NULL
  if (is.list(distortion_raw)) {
    distortion_dt <- as.data.table(data.frame(
      distortion_coefficient = first_numeric_or_na(distortion_raw[["Distortion Coefficient"]]),
      pval = first_numeric_or_na(distortion_raw[["P-value"]]),
      stringsAsFactors = FALSE
    ))
    outlier_idx <- distortion_raw[["Outliers Indices"]]
    if (!is.null(outlier_idx)) {
      distortion_dt[, outliers_indices := paste(as.character(outlier_idx), collapse = ",")]
    }
  } else if (is.data.frame(distortion_raw)) {
    distortion_dt <- as.data.table(distortion_raw)
  }

  add_meta <- function(x) {
    if (is.null(x)) return(NULL)
    x[, `:=`(protein = protein_of_interest, nsnp_mr_presso = nrow(dt))]
    x
  }

  list(
    main = add_meta(main_dt),
    global = add_meta(global_dt),
    outlier = add_meta(outlier_dt),
    distortion = add_meta(distortion_dt)
  )
}

# Saves run reproducibility files to outdir_base:
#   _run_config.tsv       — all cfg settings
#   _protein_settings.tsv — protein/chromosome/position table
#   _package_versions.tsv — versions of all used packages
#   _sessionInfo.txt      — full R session info
write_run_metadata <- function(cfg, outdir_base, protein_settings) {
  cfg_dt <- data.table(
    setting = names(cfg),
    value = vapply(cfg, function(x) paste(as.character(x), collapse = ","), character(1))
  )
  safe_write_tsv(cfg_dt, file.path(outdir_base, "_run_config.tsv"))
  safe_write_tsv(as.data.table(protein_settings), file.path(outdir_base, "_protein_settings.tsv"))

  pkgs <- c("data.table", "dplyr", "TwoSampleMR", "ggplot2", "coloc", "MRPRESSO")
  pkg_dt <- data.table(
    package = pkgs,
    version = vapply(
      pkgs,
      function(p) {
        if (requireNamespace(p, quietly = TRUE)) as.character(packageVersion(p)) else NA_character_
      },
      character(1)
    )
  )
  safe_write_tsv(pkg_dt, file.path(outdir_base, "_package_versions.tsv"))

  session_txt <- capture.output(sessionInfo())
  writeLines(session_txt, con = file.path(outdir_base, "_sessionInfo.txt"))
}

# Draws a forest plot of MR estimates across methods (IVW, Egger, median, mode).
# plot_on="or" exponentiates betas to odds ratios; plot_on="beta" plots on the
# original scale. CIs that exceed the axis are shown as arrows.
make_mr_methods_forest_caption <- function(mr_res,
                                           title,
                                           outfile,
                                           sig_level = 0.05,
                                           plot_on = c("or", "beta"),
                                           auto_xlim = TRUE,
                                           xlim_exclude = c("MR Egger"),
                                           pad_mult = 1.15,
                                           x_min = NULL,
                                           x_max = NULL,
                                           xlab = "OR and 95% CI",
                                           caption = NULL) {
  plot_on <- match.arg(plot_on)

  dd <- mr_res %>%
    dplyr::select(method, b, se, pval, nsnp) %>%
    dplyr::mutate(
      method = as.character(method),
      b = as.numeric(b),
      se = as.numeric(se),
      pval = suppressWarnings(as.numeric(pval))
    ) %>%
    dplyr::filter(is.finite(b), is.finite(se))

  if (nrow(dd) == 0) stop("No finite b/se in mr_res.")

  if (plot_on == "or") {
    dd <- dd %>% mutate(est = exp(b), lo = exp(b - 1.96 * se), hi = exp(b + 1.96 * se), null = 1)
  } else {
    dd <- dd %>% mutate(est = b, lo = b - 1.96 * se, hi = b + 1.96 * se, null = 0)
  }

  preferred <- c("Inverse variance weighted", "MR Egger", "Weighted median", "Weighted mode")
  present <- unique(dd$method)
  levs <- c(intersect(preferred, present), setdiff(present, preferred))

  dd <- dd %>% mutate(method = factor(method, levels = levs)) %>% arrange(method)

  if (auto_xlim) {
    dd_lim <- dd %>% filter(!as.character(method) %in% xlim_exclude)
    if (nrow(dd_lim) == 0) dd_lim <- dd

    if (plot_on == "or") {
      xmin0 <- min(dd_lim$lo, dd$null, na.rm = TRUE)
      xmax0 <- max(dd_lim$hi, dd$null, na.rm = TRUE)
      xmin0 <- max(xmin0, 1e-12)
      x_min <- xmin0 / pad_mult
      x_max <- xmax0 * pad_mult
    } else {
      xmin0 <- min(dd_lim$lo, dd$null, na.rm = TRUE)
      xmax0 <- max(dd_lim$hi, dd$null, na.rm = TRUE)
      rng <- xmax0 - xmin0
      x_min <- xmin0 - 0.05 * rng
      x_max <- xmax0 + 0.05 * rng
    }
  } else if (is.null(x_min) || is.null(x_max)) {
    stop("If auto_xlim=FALSE, provide x_min and x_max.")
  }

  dd <- dd %>%
    mutate(
      sig = ifelse(!is.na(pval) & pval < sig_level, "sig", "ns"),
      lo_clip = pmax(lo, x_min),
      hi_clip = pmin(hi, x_max),
      right_over = hi > x_max,
      left_over = lo < x_min
    )

  if (plot_on == "or") {
    dd <- dd %>% mutate(est_plot = pmin(pmax(est, x_min * 1.02), x_max / 1.02))
  } else {
    eps <- 0.01 * (x_max - x_min)
    dd <- dd %>% mutate(est_plot = pmin(pmax(est, x_min + eps), x_max - eps))
  }

  p <- ggplot(dd, aes(y = method, x = est_plot)) +
    geom_vline(xintercept = unique(dd$null), linetype = "dashed") +
    geom_errorbarh(aes(xmin = lo_clip, xmax = hi_clip, color = sig), height = 0) +
    geom_point(aes(color = sig), size = 3) +
    geom_segment(data = subset(dd, right_over),
                 aes(x = x_max - 0.03 * (x_max - x_min), xend = x_max, y = method, yend = method, color = sig),
                 arrow = arrow(length = unit(0.12, "inches"))) +
    geom_segment(data = subset(dd, left_over),
                 aes(x = x_min + 0.03 * (x_max - x_min), xend = x_min, y = method, yend = method, color = sig),
                 arrow = arrow(length = unit(0.12, "inches"))) +
    scale_color_manual(values = c(sig = "red", ns = "black")) +
    coord_cartesian(xlim = c(x_min, x_max)) +
    scale_y_discrete(limits = rev(levs)) +
    labs(title = title, x = xlab, y = NULL, caption = caption) +
    theme_classic() +
    theme(legend.position = "none", plot.caption = element_text(hjust = 0))

  ggsave(outfile, plot = p, width = 6, height = 4, dpi = 200)
  p
}
```

## Step 1: Read and Standardize Inputs

```{r step1_read_inputs}
exp0 <- fread(cfg$exposure_path, check.names = FALSE)
out0 <- fread(cfg$outcome_path, check.names = FALSE)
bim <- read_bim(cfg$ld_bfile)

ld_fix <- ensure_unique_ld_bfile(
  bfile = cfg$ld_bfile,
  bim = bim,
  outdir_base = cfg$outdir_base,
  force = FALSE
)
ld_bfile <- ld_fix$bfile
bim <- ld_fix$bim

exp1 <- standardise_exposure(exp0)
out1 <- standardise_outcome(out0)
out1_best <- deduplicate_by_snp(out1[is.finite(beta) & is.finite(se) & se > 0])

setindexv(exp1, c("protein_key", "CHR", "POS", "snp"))
setindexv(out1, c("snp"))
setindexv(out1, c("CHR", "POS"))
setindexv(out1_best, c("snp"))
setindexv(out1_best, c("CHR", "POS"))
setindexv(bim, c("SNP"))

cat("Exposure rows raw:", nrow(exp0), "| standardized:", nrow(exp1), "\n")
cat("Outcome rows raw:", nrow(out0), "| standardized:", nrow(out1), "\n")
cat("Outcome rows for lookup:", nrow(out1_best), "\n")
cat("LD BIM rows:", nrow(bim), "\n")

safe_write_tsv(exp1[, !c("protein_key"), with = FALSE], file.path(cfg$outdir_base, "_global_exposure_standardized.tsv"))
safe_write_tsv(out1, file.path(cfg$outdir_base, "_global_outcome_standardized.tsv"))
write_run_metadata(cfg, cfg$outdir_base, protein_settings)
```

## Step 2: Per-Protein MR Function

```{r step2_function}
run_one_protein <- function(protein_of_interest, gene_chr, gene_pos, exp1, out1, out1_best, bim, ld_bfile, cfg) {
  exposure_name <- protein_of_interest
  outdir <- file.path(cfg$outdir_base, protein_of_interest)
  dir.create(outdir, showWarnings = FALSE, recursive = TRUE)

  diag <- instrument_diag_init(protein_of_interest, gene_chr, gene_pos, cfg)
  diag_fail_step <- NA_character_
  diag_fail_message <- NA_character_
  nsnp <- NA_integer_

  mr_res <- NULL
  res_wald <- NULL
  res_wald_top <- NULL
  het <- NULL
  pleio <- NULL
  loo <- NULL
  loo_s <- NULL
  mr_out <- NULL
  singlesnp_res <- NULL
  steiger_summary <- NULL
  mr_presso_main <- NULL
  mr_presso_global <- NULL
  mr_presso_outlier <- NULL
  mr_presso_distortion <- NULL
  coloc_summary <- NULL
  optional_failures <- data.table(step = character(), message = character())
  protein_key_value <- toupper(protein_of_interest)

  tryCatch({
    # Define the cis window: gene position ± cfg$cis_kb kilobases.
    cis_start <- gene_pos - cfg$cis_kb * 1000
    cis_end <- gene_pos + cfg$cis_kb * 1000
    log10p_thresh <- -log10(cfg$p_threshold)

    # Step 1: Subset exposure SNPs to cis window for this protein.
    exp_prot_all <- exp1[protein_key == protein_key_value]
    exp_chr_all <- exp_prot_all[CHR == gene_chr]
    exp_cis_all <- exp_chr_all[POS >= cis_start & POS <= cis_end]

    diag <- instrument_diag_set(
      diag,
      n_exp_protein_total = nrow(exp_prot_all),
      n_exp_chr_total = nrow(exp_chr_all),
      n_exp_cis_total = nrow(exp_cis_all)
    )

    cat("----\n")
    cat("Protein:", protein_of_interest, "| chr:", gene_chr, "| pos:", gene_pos, "\n")

    # Step 2: Apply genome-wide significance threshold (P < cfg$p_threshold).
    exp2 <- copy(exp_cis_all[neglog10p >= log10p_thresh])

    diag <- instrument_diag_set(
      diag,
      n_exp_after_p = nrow(exp2),
      n_exp_with_eaf = sum(is.finite(exp2$eaf)),
    )
    if (nrow(exp2) == 0) {
      diag_fail_step <- "cis+pval"
      diag_fail_message <- "No exposure variants after cis + p-value filtering."
      stop(diag_fail_message)
    }

    # Step 3: F-statistic filter for instrument strength (F = (beta/se)^2 ≥ 10).
    exp2[, Fstat := (beta / se)^2]
    exp2 <- exp2[Fstat >= cfg$f_threshold]
    diag <- instrument_diag_set(diag, n_exp_after_f = nrow(exp2))
    if (nrow(exp2) == 0) {
      diag_fail_step <- "Fstat"
      diag_fail_message <- "All exposure instruments removed by F-stat filter."
      stop(diag_fail_message)
    }

    exp2 <- deduplicate_by_snp(exp2)
    diag <- instrument_diag_set(diag, n_exp_unique_snp = nrow(exp2))

    cat("Exposure instruments before LD overlap:", nrow(exp2), "\n")

    # Step 4: Split instruments into those present in the LD reference and those
    # absent. Only SNPs in the LD reference can be LD-clumped; missing ones are
    # saved for review but cannot be used as instruments.
    bim_ids <- bim$SNP
    exp2_in_ld <- copy(exp2[snp %chin% bim_ids])
    exp2_missing <- copy(exp2[!snp %chin% bim_ids])

    diag <- instrument_diag_set(
      diag,
      n_exp_in_ld = nrow(exp2_in_ld),
      n_exp_missing_ld = nrow(exp2_missing),
      n_exp_to_clump = nrow(exp2_in_ld)
    )

    cat("LD overlap:", nrow(exp2_in_ld), "of", nrow(exp2), "\n")

    if (nrow(exp2_in_ld) == 0) {
      diag_fail_step <- "ld_overlap"
      diag_fail_message <- "0 overlap between exposure instruments and LD reference. Check build or variant ID mapping."
      stop(diag_fail_message)
    }

    # Step 5: LD clumping to select independent instruments (r² < cfg$clump_r2).
    # Skipped when only one SNP is present in the LD reference.
    if (nrow(exp2_in_ld) == 1) {
      exp3 <- copy(exp2_in_ld)
      cat("Only one instrument in LD reference; clumping skipped.\n")
    } else {
      exp3 <- clump_exposure_local_tsmr(
        exp_df = exp2_in_ld,
        cfg = cfg,
        ld_bfile = ld_bfile,
        protein_of_interest = protein_of_interest,
        outdir = outdir
      )
    }

    exp3 <- deduplicate_by_snp(exp3)
    diag <- instrument_diag_set(diag, n_exp_after_clump = nrow(exp3))

    safe_write_tsv(exp3, file.path(outdir, "exposure_instruments_final.tsv"))
    if (nrow(exp2_missing) > 0) safe_write_tsv(exp2_missing, file.path(outdir, "instruments_missing_in_ld.tsv"))

    target_snps <- unique(exp3$snp)
    out_use <- copy(out1_best[snp %chin% target_snps])

    missing <- setdiff(target_snps, unique(out_use$snp))
    if (length(missing) > 0) {
      safe_write_tsv(
        data.table(missing_snp = missing),
        file.path(outdir, paste0(protein_of_interest, "_missing_in_outcome_snps.tsv"))
      )
    }

    # Step 6: Match instruments to outcome. For SNPs absent from the outcome
    # GWAS, search for LD proxies (r² ≥ cfg$proxy_r2_min) and substitute them.
    if (isTRUE(cfg$use_outcome_proxies)) {
      proxy_res <- rescue_outcome_proxies(
        target_snps = target_snps,
        exp_df = exp3,
        out_best = out1_best,
        bim = bim,
        cfg = cfg,
        ld_bfile = ld_bfile
      )

      safe_write_tsv(proxy_res$search_results, file.path(outdir, "proxy_search_results.tsv"))
      safe_write_tsv(proxy_res$status, file.path(outdir, "proxy_substitutions.tsv"))
      safe_write_tsv(proxy_res$unresolved, file.path(outdir, "missing_in_outcome_no_proxy.tsv"))

      rescued_snps <- unique(proxy_res$status[status == "rescued_by_proxy", target_snp])
      keep_snps <- unique(proxy_res$status[status %in% c("found_natively", "rescued_by_proxy"), target_snp])

      exp3 <- exp3[snp %chin% keep_snps]
      target_snps <- unique(exp3$snp)
      out_use <- copy(out1_best[snp %chin% target_snps])
      if (nrow(proxy_res$rescued_rows) > 0) {
        out_use <- rbindlist(list(out_use, proxy_res$rescued_rows), fill = TRUE, use.names = TRUE)
        out_use <- out_use[!(snp %chin% rescued_snps & (is.na(proxy_source) | proxy_source != "local_plink_ld"))]
      }
    } else if (length(missing) > 0) {
      exp3 <- exp3[!snp %chin% missing]
      target_snps <- unique(exp3$snp)
      out_use <- out_use[!snp %chin% missing]
    }

    out_use <- deduplicate_by_snp(out_use)

    diag <- instrument_diag_set(
      diag,
      n_outcome_overlap = nrow(out_use),
      n_outcome_with_eaf = sum(is.finite(out_use$eaf)),
    )

    if (length(target_snps) == 0 || nrow(exp3) == 0 || nrow(out_use) == 0) {
      diag_fail_step <- "outcome_overlap"
      diag_fail_message <- "No instruments remain after outcome overlap filtering."
      stop(diag_fail_message)
    }

    exposure_dat <- format_tsmr_data(exp3, type = "exposure", label = exposure_name)
    outcome_dat <- format_tsmr_data(out_use, type = "outcome", label = cfg$outcome_name)

    common <- intersect(exposure_dat$SNP, outcome_dat$SNP)
    exposure_dat <- exposure_dat[exposure_dat$SNP %in% common, ]
    outcome_dat <- outcome_dat[outcome_dat$SNP %in% common, ]

    diag <- instrument_diag_set(diag, n_after_format_overlap = length(common))

    if (nrow(exposure_dat) == 0) {
      diag_fail_step <- "format_overlap"
      diag_fail_message <- "No overlapping SNPs after formatting."
      stop(diag_fail_message)
    }

    # Step 7: Harmonise alleles between exposure and outcome. action=2 tries to
    # align ambiguous palindromic SNPs using EAF; drop_palindromic_harmonised()
    # then removes all palindromics regardless (conservative, equivalent to action=3).
    dat_h <- harmonise_data(exposure_dat, outcome_dat, action = cfg$harmonise_action)
    dat_h <- drop_palindromic_harmonised(dat_h)

    audit <- attr(dat_h, "audit")
    if (!is.null(audit)) {
      safe_write_tsv(as.data.table(audit), file.path(outdir, "harmonise_audit.tsv"))
    }

    pal_removed <- attr(dat_h, "palindromic_removed")
    if (!is.null(pal_removed) && nrow(pal_removed) > 0) {
      safe_write_tsv(as.data.table(pal_removed), file.path(outdir, "palindromic_snps_removed.tsv"))
    }

    d1 <- dat_h[dat_h$mr_keep == TRUE, ]
    nsnp <- nrow(d1)
    diag <- instrument_diag_set(
      diag,
      n_after_harmonise = nrow(dat_h),
      n_palindromic_removed = if (!is.null(pal_removed)) nrow(as.data.table(pal_removed)) else 0L,
      n_mr_keep = nsnp
    )
    if (nsnp == 0) {
      diag_fail_step <- "harmonise_mr_keep"
      diag_fail_message <- "No valid instruments after harmonisation."
      stop(diag_fail_message)
    }

    # Step 8: Run MR. With 1 instrument: Wald ratio. With ≥2: IVW (primary);
    # weighted median added at ≥3 SNPs; Egger and weighted mode at ≥10 SNPs.
    # A "top SNP" Wald ratio is always computed alongside multi-instrument IVW
    # for comparison.
    if (nsnp == 1) {
      bx <- d1$beta.exposure
      bxse <- d1$se.exposure
      by <- d1$beta.outcome
      byse <- d1$se.outcome

      if (abs(bx) < 1e-8) stop("Exposure beta too close to 0 for Wald ratio.")

      b_wald <- by / bx
      se_wald <- sqrt((byse^2) / (bx^2) + (by^2 * bxse^2) / (bx^4))
      p_wald <- 2 * pnorm(abs(b_wald / se_wald), lower.tail = FALSE)

      res_wald <- data.frame(
        protein = protein_of_interest,
        method = "Wald ratio",
        nsnp = 1,
        b = b_wald,
        se = se_wald,
        pval = p_wald,
        top_snp = as.character(d1$SNP),
        stringsAsFactors = FALSE
      )

      mr_res <- res_wald
    } else {
      methods <- c("mr_ivw")
      if (nsnp >= 3) methods <- c(methods, "mr_weighted_median")
      if (nsnp >= 10) methods <- c(methods, "mr_egger_regression", "mr_weighted_mode")

      mr_out <- mr(d1, method_list = methods)
      mr_res <- mr_out %>%
        dplyr::mutate(protein = protein_of_interest, top_snp = NA_character_) %>%
        dplyr::select(protein, method, nsnp, b, se, pval, top_snp) %>%
        as.data.frame()

      d1_dt <- as.data.table(d1)
      d1_dt[, pval.exposure := suppressWarnings(as.numeric(pval.exposure))]
      if (all(!is.finite(d1_dt$pval.exposure))) {
        d1_dt[, pval.exposure := 2 * pnorm(-abs(beta.exposure / se.exposure))]
      }

      d1_dt <- d1_dt[is.finite(pval.exposure)]
      if (nrow(d1_dt) > 0) {
        d1_dt[, Fstat := (beta.exposure / se.exposure)^2]
        d1_dt[, abs_bx := abs(beta.exposure)]
        setorder(d1_dt, pval.exposure, -Fstat, -abs_bx)
        top_snp <- d1_dt[1]

        b_top <- top_snp$beta.outcome / top_snp$beta.exposure
        se_top <- sqrt((top_snp$se.outcome^2) / (top_snp$beta.exposure^2) +
                       (top_snp$beta.outcome^2 * top_snp$se.exposure^2) / (top_snp$beta.exposure^4))
        p_top <- 2 * pnorm(abs(b_top / se_top), lower.tail = FALSE)

        res_wald_top <- data.frame(
          protein = protein_of_interest,
          method = "Wald ratio (top SNP)",
          nsnp = 1,
          b = b_top,
          se = se_top,
          pval = p_top,
          top_snp = top_snp$SNP,
          stringsAsFactors = FALSE
        )

        mr_res <- bind_rows(mr_res, res_wald_top) %>% as.data.frame()
      }

      if (nsnp >= 3) {
        het <- mr_heterogeneity(d1)
        pleio <- mr_pleiotropy_test(d1)
      }

      loo <- mr_leaveoneout(d1)
      loo_s <- loo %>%
        mutate(
          exp_beta = exp(b),
          exp_beta_lci95 = exp(b - 1.96 * se),
          exp_beta_uci95 = exp(b + 1.96 * se)
        ) %>%
        arrange(p)
    }

    safe_write_tsv(dat_h, file.path(outdir, "harmonised_dat.tsv"))
    if (!is.null(mr_res)) safe_write_tsv(mr_res, file.path(outdir, "mr_results.tsv"))
    if (!is.null(res_wald_top)) safe_write_tsv(res_wald_top, file.path(outdir, "wald_ratio_top_snp.tsv"))
    if (!is.null(het)) safe_write_tsv(het, file.path(outdir, "heterogeneity.tsv"))
    if (!is.null(pleio)) safe_write_tsv(pleio, file.path(outdir, "pleiotropy.tsv"))
    if (!is.null(loo)) safe_write_tsv(loo, file.path(outdir, "leaveoneout.tsv"))
    if (!is.null(loo_s)) safe_write_tsv(loo_s, file.path(outdir, "leaveoneout_summary.tsv"))

    if (isTRUE(cfg$run_steiger)) {
      tryCatch({
        steiger_summary <- run_steiger_directionality(d1, cfg, protein_of_interest)
        if (!is.null(steiger_summary)) {
          safe_write_tsv(steiger_summary, file.path(outdir, "steiger_summary.tsv"))
        } else {
          cat("Steiger skipped for", protein_of_interest, ": missing p/N information after harmonization.\n")
        }
      }, error = function(e) {
        optional_failures <<- rbind(
          optional_failures,
          data.table(step = "Steiger", message = e$message),
          fill = TRUE
        )
        message("Optional step failed for ", protein_of_interest, " [Steiger]: ", e$message)
      })
    }

    if (isTRUE(cfg$run_mr_presso)) {
      tryCatch({
        presso <- run_mr_presso_test(d1, cfg, protein_of_interest)
        if (!is.null(presso)) {
          mr_presso_main <- presso$main
          mr_presso_global <- presso$global
          mr_presso_outlier <- presso$outlier
          mr_presso_distortion <- presso$distortion

          if (!is.null(mr_presso_main)) safe_write_tsv(mr_presso_main, file.path(outdir, "mr_presso_main.tsv"))
          if (!is.null(mr_presso_global)) safe_write_tsv(mr_presso_global, file.path(outdir, "mr_presso_global.tsv"))
          if (!is.null(mr_presso_outlier)) safe_write_tsv(mr_presso_outlier, file.path(outdir, "mr_presso_outlier.tsv"))
          if (!is.null(mr_presso_distortion)) safe_write_tsv(mr_presso_distortion, file.path(outdir, "mr_presso_distortion.tsv"))
        } else {
          cat("MR-PRESSO skipped for", protein_of_interest, ": requires >= ", cfg$mr_presso_min_snps, " valid SNPs.\n")
        }
      }, error = function(e) {
        optional_failures <<- rbind(
          optional_failures,
          data.table(step = "MR-PRESSO", message = e$message),
          fill = TRUE
        )
        message("Optional step failed for ", protein_of_interest, " [MR-PRESSO]: ", e$message)
      })
    }

    if (isTRUE(cfg$run_coloc)) {
      tryCatch({
        exp_reg <- subset_exposure_region(exp1, protein_key_value, gene_chr, cis_start, cis_end)
        out_reg <- subset_outcome_region(out1_best, gene_chr, cis_start, cis_end)

        if (nrow(exp_reg) > 0 && nrow(out_reg) > 0) {
          exp_reg <- exp_reg[
            is.finite(beta) & is.finite(se) & se > 0 &
              is.finite(eaf) & eaf > 0 & eaf < 1 &
              is_acgt_snp(effect_allele, other_allele)
          ]
          out_reg <- out_reg[
            is.finite(beta) & is.finite(se) & se > 0 &
              is.finite(eaf) & eaf > 0 & eaf < 1 &
              is_acgt_snp(effect_allele, other_allele)
          ]

          exp_reg <- deduplicate_by_snp(exp_reg)
          out_reg <- deduplicate_by_snp(out_reg)

          common_reg <- intersect(exp_reg$snp, out_reg$snp)
          if (length(common_reg) >= 10L) {
            exp_reg <- exp_reg[snp %chin% common_reg]
            out_reg <- out_reg[snp %chin% common_reg]

            exp_reg_dat <- format_tsmr_data(exp_reg, type = "exposure", label = exposure_name)
            out_reg_dat <- format_tsmr_data(out_reg, type = "outcome", label = cfg$outcome_name)

            reg_h <- harmonise_data(exp_reg_dat, out_reg_dat, action = cfg$harmonise_action)
            reg_h <- drop_palindromic_harmonised(reg_h)
            reg_h <- reg_h[reg_h$mr_keep == TRUE, ]

            if (nrow(reg_h) >= 10L) {
              D1 <- list(
                type = "quant",
                snp = reg_h$SNP,
                beta = reg_h$beta.exposure,
                varbeta = (reg_h$se.exposure)^2,
                MAF = pmin(reg_h$eaf.exposure, 1 - reg_h$eaf.exposure)
              )

              if ("samplesize.exposure" %in% names(reg_h)) {
                N1 <- suppressWarnings(as.numeric(reg_h$samplesize.exposure))
                if (any(is.finite(N1))) D1$N <- round(median(N1[is.finite(N1)]))
              }

              D2 <- list(
                type = cfg$outcome_type_coloc,
                snp = reg_h$SNP,
                beta = reg_h$beta.outcome,
                varbeta = (reg_h$se.outcome)^2,
                MAF = pmin(reg_h$eaf.outcome, 1 - reg_h$eaf.outcome)
              )

              if (is.finite(cfg$outcome_N)) D2$N <- cfg$outcome_N
              if (identical(cfg$outcome_type_coloc, "cc") && is.finite(cfg$outcome_s)) D2$s <- cfg$outcome_s

              coloc_res <- NULL
              if (identical(cfg$outcome_type_coloc, "cc") && !is.finite(cfg$outcome_s)) {
                s_grid <- c(0.05, 0.10, 0.20)
                coloc_list <- lapply(s_grid, function(s) {
                  D2i <- D2
                  D2i$s <- s
                  coloc.abf(D1, D2i)
                })

                coloc_sum <- rbindlist(lapply(seq_along(coloc_list), function(i) {
                  data.table(s = s_grid[i], t(coloc_list[[i]]$summary))
                }), fill = TRUE)

                safe_write_tsv(coloc_sum, file.path(outdir, paste0(protein_of_interest, "_coloc_summary_sgrid.tsv")))
                coloc_res <- coloc_list[[2]]
              } else {
                coloc_res <- coloc.abf(D1, D2)
              }

              coloc_summary <- as.data.table(t(coloc_res$summary))
              coloc_summary[, `:=`(
                protein = protein_of_interest,
                cis_kb = cfg$cis_kb,
                chr = gene_chr,
                start = cis_start,
                end = cis_end
              )]

              coloc_snps <- as.data.table(coloc_res$results)
              coloc_snps[, protein := protein_of_interest]

              safe_write_tsv(coloc_summary, file.path(outdir, paste0(protein_of_interest, "_coloc_summary.tsv")))
              safe_write_tsv(coloc_snps, file.path(outdir, paste0(protein_of_interest, "_coloc_snps.tsv")))
            }
          }
        }
      }, error = function(e) {
        optional_failures <<- rbind(
          optional_failures,
          data.table(step = "Coloc", message = e$message),
          fill = TRUE
        )
        message("Optional step failed for ", protein_of_interest, " [Coloc]: ", e$message)
      })
    }

    if (!is.null(mr_res) && nrow(d1) > 1) {
      tryCatch({
        p_scatter <- mr_scatter_plot(mr_out, d1)
        ggsave(file.path(outdir, "mr_scatter.png"), p_scatter[[1]], width = 6, height = 5, dpi = 200)
      }, error = function(e) {
        optional_failures <<- rbind(
          optional_failures,
          data.table(step = "MR scatter plot", message = e$message),
          fill = TRUE
        )
        message("Optional step failed for ", protein_of_interest, " [MR scatter plot]: ", e$message)
      })

      tryCatch({
        make_mr_methods_forest_caption(
          mr_res = mr_res,
          title = exposure_name,
          outfile = file.path(outdir, paste0("mr_forest_methods_", protein_of_interest, ".png")),
          plot_on = "or",
          auto_xlim = TRUE,
          xlim_exclude = c("MR Egger"),
          caption = paste0("Forest plot of MR OR estimates for ", exposure_name, ".")
        )
      }, error = function(e) {
        message("Methods forest plot skipped for ", protein_of_interest, ": ", e$message)
      })
    }

    tryCatch({
      singlesnp_res <- mr_singlesnp(dat_h)
      p_forest <- mr_forest_plot(singlesnp_res)
      ggsave(file.path(outdir, "mr_forest_singlesnp.png"), p_forest[[1]], width = 7, height = 7, dpi = 200)
    }, error = function(e) {
      optional_failures <<- rbind(
        optional_failures,
        data.table(step = "Single-SNP forest plot", message = e$message),
        fill = TRUE
      )
      message("Optional step failed for ", protein_of_interest, " [Single-SNP forest plot]: ", e$message)
    })

    if (!is.null(loo)) {
      tryCatch({
        p_loo <- mr_leaveoneout_plot(loo)
        ggsave(file.path(outdir, "mr_leaveoneout.png"), p_loo[[1]], width = 7, height = 6, dpi = 200)
      }, error = function(e) {
        optional_failures <<- rbind(
          optional_failures,
          data.table(step = "Leave-one-out plot", message = e$message),
          fill = TRUE
        )
        message("Optional step failed for ", protein_of_interest, " [Leave-one-out plot]: ", e$message)
      })
    }

    if (nrow(optional_failures) > 0) {
      safe_write_tsv(optional_failures, file.path(outdir, "optional_step_failures.tsv"))
    }

    list(
      protein = protein_of_interest,
      ok = !is.null(mr_res) && nrow(mr_res) > 0,
      nsnp = nsnp,
      outdir = outdir,
      mr_table = mr_res,
      steiger_summary = steiger_summary,
      mr_presso_main = mr_presso_main,
      mr_presso_global = mr_presso_global,
      coloc_summary = coloc_summary,
      diag = instrument_diag_complete(diag, status_value = "ok")
    )
  }, error = function(e) {
    if (is.na(diag_fail_step)) diag_fail_step <<- "unhandled"
    if (is.na(diag_fail_message) || !nzchar(diag_fail_message)) {
      diag_fail_message <<- conditionMessage(e)
    }
    message("FAILED for ", protein_of_interest, ": ", conditionMessage(e))

    list(
      protein = protein_of_interest,
      ok = FALSE,
      error = conditionMessage(e),
      nsnp = nsnp,
      outdir = outdir,
      mr_table = NULL,
      steiger_summary = NULL,
      mr_presso_main = NULL,
      mr_presso_global = NULL,
      coloc_summary = NULL,
      diag = instrument_diag_complete(
        diag,
        status_value = "failed",
        fail_step_value = diag_fail_step,
        fail_message_value = diag_fail_message
      )
    )
  })
}
```

## Step 3: Run All Proteins

```{r step3_run_all}
run_one_protein_index <- function(i) {
  run_one_protein(
    protein_of_interest = protein_settings$protein[[i]],
    gene_chr = protein_settings$gene_chr[[i]],
    gene_pos = protein_settings$gene_pos[[i]],
    exp1 = exp1,
    out1 = out1,
    out1_best = out1_best,
    bim = bim,
    ld_bfile = ld_bfile,
    cfg = cfg
  )
}

if (cfg$n_workers > 1L && .Platform$OS.type == "unix") {
  results <- parallel::mclapply(
    seq_len(nrow(protein_settings)),
    run_one_protein_index,
    mc.cores = cfg$n_workers
  )
} else {
  results <- lapply(seq_len(nrow(protein_settings)), run_one_protein_index)
}

names(results) <- protein_settings$protein
results <- results[protein_settings$protein]

diag_rows <- lapply(results, function(x) x$diag)
diag_rows <- Filter(Negate(is.null), diag_rows)
tryCatch(
  instrument_diag_write(diag_rows, cfg$outdir_base),
  error = function(e) message("Instrument audit write failed: ", e$message)
)

# -- Combined MR results --
ok_results <- results[vapply(results, function(x) isTRUE(x$ok), logical(1))]
if (length(ok_results) > 0) {
  mr_tables <- lapply(ok_results, function(x) as.data.table(x$mr_table))
  mr_all <- rbindlist(mr_tables, fill = TRUE)
  mr_all <- mr_all[order(pval)]

  mr_all[, p_fdr_by_method := p.adjust(pval, method = "fdr"), by = method]
  mr_all[, p_fdr_primary := NA_real_]
  mr_all[method %in% c("Inverse variance weighted", "Wald ratio"), p_fdr_primary := p.adjust(pval, method = "fdr")]

  safe_write_tsv(mr_all, file.path(cfg$outdir_base, "MR_all_proteins_summary.tsv"))
  mr_primary <- mr_all[method %in% c("Inverse variance weighted", "Wald ratio")]
  safe_write_tsv(mr_primary, file.path(cfg$outdir_base, "MR_primary_methods_summary.tsv"))
  print(mr_all)
} else {
  cat("No successful protein runs for MR summary.\n")
}

# -- Steiger directionality --
if (isTRUE(cfg$run_steiger)) {
  steiger_tables <- lapply(ok_results, function(x) x$steiger_summary)
  steiger_tables <- Filter(Negate(is.null), steiger_tables)

  if (length(steiger_tables) > 0) {
    steiger_all <- rbindlist(steiger_tables, fill = TRUE)
    safe_write_tsv(steiger_all, file.path(cfg$outdir_base, "Steiger_all_proteins_summary.tsv"))
    print(steiger_all)
  } else {
    cat("No Steiger summaries available to combine.\n")
  }
}

# -- MR-PRESSO results --
if (isTRUE(cfg$run_mr_presso)) {
  presso_main_tables <- lapply(ok_results, function(x) x$mr_presso_main)
  presso_main_tables <- Filter(Negate(is.null), presso_main_tables)

  if (length(presso_main_tables) > 0) {
    presso_main_all <- rbindlist(presso_main_tables, fill = TRUE)
    safe_write_tsv(presso_main_all, file.path(cfg$outdir_base, "MRPRESSO_all_proteins_main.tsv"))
    print(presso_main_all)
  } else {
    cat("No MR-PRESSO main summaries available to combine.\n")
  }

  presso_global_tables <- lapply(ok_results, function(x) x$mr_presso_global)
  presso_global_tables <- Filter(Negate(is.null), presso_global_tables)

  if (length(presso_global_tables) > 0) {
    presso_global_all <- rbindlist(presso_global_tables, fill = TRUE)
    safe_write_tsv(presso_global_all, file.path(cfg$outdir_base, "MRPRESSO_all_proteins_global.tsv"))
    print(presso_global_all)
  } else {
    cat("No MR-PRESSO global summaries available to combine.\n")
  }
}

# -- Colocalization results --
if (isTRUE(cfg$run_coloc)) {
  coloc_tables <- lapply(ok_results, function(x) x$coloc_summary)
  coloc_tables <- Filter(Negate(is.null), coloc_tables)

  if (length(coloc_tables) > 0) {
    coloc_all <- rbindlist(coloc_tables, fill = TRUE)
    safe_write_tsv(coloc_all, file.path(cfg$outdir_base, "Coloc_all_proteins_summary.tsv"))
    print(coloc_all)
  } else {
    cat("No coloc summaries available to combine.\n")
  }
}

# -- Failed proteins --
fail_results <- results[vapply(results, function(x) isFALSE(x$ok), logical(1))]
if (length(fail_results) > 0) {
  fail_dt <- rbindlist(lapply(fail_results, function(x) {
    diag_dt <- if (!is.null(x$diag)) {
      as.data.table(x$diag)[, .(fail_step, fail_message)]
    } else {
      data.table(fail_step = NA_character_, fail_message = NA_character_)
    }

    cbind(
      data.table(
        protein = x$protein,
        ok = x$ok,
        error = x$error,
        nsnp = x$nsnp,
        outdir = x$outdir
      ),
      diag_dt
    )
  }), fill = TRUE)
  safe_write_tsv(fail_dt, file.path(cfg$outdir_base, "MR_failed_runs.tsv"))
  print(fail_dt)
}

cat("Run complete at:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n")
```

## Session Info

```{r session_info}
sessionInfo()
```
